### 2015-09-11 BPLE/Manuscript.tex
first commit. empty repo

% an empty .tex file. Eventually may transwer .odt in here

### 2015-09-11 BPLE/R/lookatdata.R
some materials from dropbox

@@ -1,0 +1,40 @@
+# Author: triffe
+###############################################################################
+
+
+States <- local(get(load("/hdir/0/triffe/workspace/other/Decomposition/StateDivergence/Data/MXClong.Rdata")))
+LT     <- local(get(load("/hdir/0/triffe/workspace/other/Decomposition/StateDivergence/Data/MilaStatesSingleAge.Rdata")))
+head(LT)
+fRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "f"], LT$Year[LT$Age == 0& LT$Sex == "f"], max)
+mRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "m"], LT$Year[LT$Age == 0& LT$Sex == "m"], max)
+
+plot(unique(LT$Year), RLE, type = 'l', main = "US Record e(0)")
+States$Cause <- as.character(States$Cause)
+
+library(data.table)
+States <- data.table(States)
+Mins <- States[,min(mxc),by=list(Year, Sex, Age, Cause)]
+Agg <- Mins[,sum(V1),by=list(Year, Sex, Age)]
+
+library(reshape2)
+Males <- acast(Agg[Agg$Sex == "male",], Age~Year, value.var = "V1")
+Females <- acast(Agg[Agg$Sex == "female",], Age~Year, value.var = "V1")
+
+mx2ex <- function(mx){
+  lx <- c(1,exp(-cumsum(mx)),0)
+  Lx <- (lx[-1] + lx[-length(lx)]) / 2
+  sum(Lx)
+}
+me0 <- apply(Males, 2, mx2ex)
+fe0 <- apply(Females, 2, mx2ex)
+getwd()
+png("/hdir/0/triffe/workspace/other/Decomposition/StateDivergence/Figures/ExampleBestPractices.png")
+plot(1959:2010, fRLE, type = 'l', main = "US States\nvanguard and best practices e(0)", col = "red", lty = 2, 
+  ylim = c(65,90), xlab = "year", ylab = "e(0)")
+
+lines(1959:2010, mRLE, col = "blue", lty = 2)
+lines(1959:2004, me0,col="blue")
+lines(1959:2004, fe0,col="red")
+legend("bottomright", lty=c(1,1,2,2), col=c("red","blue","red","blue"),
+legend=c("Female Best Prac.","Male Best Prac.","Female vanguard","Male vanguard"))
+dev.off()

### 2015-09-11 BPLE/R/schematic.R

@@ -1,0 +1,23 @@
+####################
+# Adrien Remund
+
+
+# Oeppen & Vaupel 2002 (data from EDSD)
+rle <- read.table("C:/Users/REMUND/Dropbox/BPLE/Data/Jims.txt",header=T)
+pchs <- match(rle$country,levels(rle$country))
+plot(rle$year,rle$rle,ylim=c(70,90),xlim=c(1950,2000),pch=pchs,xlab="",ylab="e0",las=1,main="Expected evolution of the RLE and BPLE")
+legend("bottomright",ncol=2,legend=levels(rle$country),pch=1:8)
+
+# schematical
+library(RColorBrewer)
+cols <- brewer.pal(3,"Set1")
+mat <- matrix(NA,ncol=4,nrow=length(1900:2020))
+mat[,1] <- 1900:2020
+mat[,2] <- 60+0.25*(0:120)
+mat[,3] <- c(55+0.32*(0:55),55+0.32*55+0.23*(0:64))
+breaks <- c(1900,1950,1970,1990,2020)
+jumps <- c(0,5,3,2)
+mat[1,4] <- 70
+for(i in 2:5){mat[mat[,1] %in% breaks[i-1]:breaks[i],4] <- mat[mat[,1] == breaks[i-1],4]+jumps[i-1]+(breaks[i-1]:breaks[i])*0.1-breaks[i-1]*0.1}
+matplot(mat[,1],mat[,2:4],las=1,ylim=c(50,100),col=cols,lty=1,lwd=2,type="s",xlab="time",ylab="e0",main="schematic evolution of life expectancy")
+legend("bottomright",legend=c("Oeppen & Vaupel 2002","Vallin & Meslé 2010","BPLE"),col=cols,lwd=2)

### 2015-09-12 BPLE/R/lookatdata.R
.

@@ -1,19 +1,37 @@
 # Author: triffe
 ###############################################################################
+# compiles long-format all-cause complete lifetables from HMD.
+# should only run once at beginning and again once at end 
+# (for final data version in paper, to be saved with metadata)
+# for Tim, this will choke
 
+hm<- local(get(load("/home/tim/Desktop/Ken-decompos/mxcod.Rdata")))
+dimnames(hm)
 
-States <- local(get(load("/hdir/0/triffe/workspace/other/Decomposition/StateDivergence/Data/MXClong.Rdata")))
-LT     <- local(get(load("/hdir/0/triffe/workspace/other/Decomposition/StateDivergence/Data/MilaStatesSingleAge.Rdata")))
-head(LT)
+if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
+	# if I'm on the laptop
+	setwd("/home/tim/git/BPLE/BPLE")
+} else {
+	# in that case I'm on Berkeley system, and other people in the dept can run this too
+	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
+}
+getwd()
+length(unique())
+States <- local(get(load("Data/MXClong.Rdata")))
+LT     <- local(get(load("Data/MilaStatesSingleAge.Rdata")))
+
 fRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "f"], LT$Year[LT$Age == 0& LT$Sex == "f"], max)
 mRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "m"], LT$Year[LT$Age == 0& LT$Sex == "m"], max)
 
-plot(unique(LT$Year), RLE, type = 'l', main = "US Record e(0)")
+plot(unique(LT$Year), fRLE, type = 'l', main = "US Record e(0)",ylim=c(65,85))
+lines(unique(LT$Year), mRLE)
 States$Cause <- as.character(States$Cause)
+unique(States$Cause)
+
 
 library(data.table)
 States <- data.table(States)
-Mins <- States[,min(mxc),by=list(Year, Sex, Age, Cause)]
+Mins   <- States[,min(mxc),by=list(Year, Sex, Age, Cause)]
 Agg <- Mins[,sum(V1),by=list(Year, Sex, Age)]
 
 library(reshape2)

### 2015-09-19 BPLE/R/DataPrep.R
got results, remade graphic

@@ -150,7 +150,9 @@ LTC[,prop2 := makeprop(mxcs), by = list(Year, State, Sex, Age)]
 LTC$mxcs2 <- LTC$prop2 * LTC$mxs
 
 # take look at example for small state...
-plot(with(LTC, mxcs2[Sex == "f" & Year == 1959 & Cause == "Cardio."]), log = 'y', type = 'l')
+#plot(with(LTC, mxcs2[Sex == "f" & Year == 1959 & Cause == "Breast"& State == "AK"]), log = 'y', type = 'l')
+#points(with(LTC, mxcs[Sex == "f" & Year == 1959 & Cause == "Breast"& State == "AK"]))
+#points(with(LTC, mxc5[Sex == "f" & Year == 1959 & Cause == "Breast"& State == "AK"]),pch=3)
 
 # so depending on whether the result is smooth or not, may be due for a second round of smoothing
 # within cause. Or maybe we accept this? Probably enough juice for now.

### 2015-09-19 BPLE/R/lookatdata.R

@@ -5,8 +5,8 @@
 # (for final data version in paper, to be saved with metadata)
 # for Tim, this will choke
 
-hm<- local(get(load("/home/tim/Desktop/Ken-decompos/mxcod.Rdata")))
-dimnames(hm)
+#hm<- local(get(load("/home/tim/Desktop/Ken-decompos/mxcod.Rdata")))
+#dimnames(hm)
 
 if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
 	# if I'm on the laptop
@@ -16,8 +16,8 @@ if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
 	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
 }
 getwd()
-length(unique())
-States <- local(get(load("Data/MXClong.Rdata")))
+
+States <- local(get(load("Data/LTC1.Rdata")))
 LT     <- local(get(load("Data/MilaStatesSingleAge.Rdata")))
 
 fRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "f"], LT$Year[LT$Age == 0& LT$Sex == "f"], max)
@@ -25,18 +25,15 @@ mRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "m"], LT$Year[LT$Age == 0& LT$Sex =
 
 plot(unique(LT$Year), fRLE, type = 'l', main = "US Record e(0)",ylim=c(65,85))
 lines(unique(LT$Year), mRLE)
-States$Cause <- as.character(States$Cause)
-unique(States$Cause)
-
 
 library(data.table)
 States <- data.table(States)
-Mins   <- States[,min(mxc),by=list(Year, Sex, Age, Cause)]
+Mins   <- States[,min(mxcs2),by=list(Year, Sex, Age, Cause)]
 Agg <- Mins[,sum(V1),by=list(Year, Sex, Age)]
 
 library(reshape2)
-Males <- acast(Agg[Agg$Sex == "male",], Age~Year, value.var = "V1")
-Females <- acast(Agg[Agg$Sex == "female",], Age~Year, value.var = "V1")
+Males <- acast(Agg[Agg$Sex == "m",], Age~Year, value.var = "V1")
[…]

### 2015-09-19 BPLE/R/lookatdata.R
overview plot redone. Smoothing made big difference in BP line. Good thing we did it!

@@ -29,11 +29,16 @@ lines(unique(LT$Year), mRLE)
 library(data.table)
 States <- data.table(States)
 Mins   <- States[,min(mxcs2),by=list(Year, Sex, Age, Cause)]
+Mins2  <- States[,min(mxc5),by=list(Year, Sex, Age, Cause)]
+
 Agg <- Mins[,sum(V1),by=list(Year, Sex, Age)]
+Agg2 <- Mins2[,sum(V1),by=list(Year, Sex, Age)]
 
 library(reshape2)
 Males <- acast(Agg[Agg$Sex == "m",], Age~Year, value.var = "V1")
 Females <- acast(Agg[Agg$Sex == "f",], Age~Year, value.var = "V1")
+Males2 <- acast(Agg2[Agg$Sex == "m",], Age~Year, value.var = "V1")
+Females2 <- acast(Agg2[Agg$Sex == "f",], Age~Year, value.var = "V1")
 
 mx2ex <- function(mx){
   lx <- c(1,exp(-cumsum(mx)),0)
@@ -42,21 +47,27 @@ mx2ex <- function(mx){
 }
 me0 <- apply(Males, 2, mx2ex)
 fe0 <- apply(Females, 2, mx2ex)
+me02 <- apply(Males2, 2, mx2ex)
+fe02 <- apply(Females2, 2, mx2ex)
 
 me0st <- acast(LT[LT$Sex == "m" & LT$Age == 0, ], Year ~ State, value.var = "ex")
 fe0st <- acast(LT[LT$Sex == "f" & LT$Age == 0, ], Year ~ State, value.var = "ex")
 
 getwd()
-png("/hdir/0/triffe/workspace/other/Decomposition/StateDivergence/Figures/ExampleBestPractices.png")
+png("Figures/ExampleBestPractices.png")
 plot(1959:2010, fRLE, type = 'l', main = "US States\nvanguard and best practices e(0)", col = "red", lty = 2, 
   ylim = c(55,90), xlab = "year", ylab = "e(0)")
 lines(1959:2010, mRLE, col = "blue", lty = 2)
 lines(1959:2004, me0,col="blue",lty=1,lwd=2)
 lines(1959:2004, fe0,col="red",lty=1,lwd=2)
+lines(1959:2004, me02,col="blue",lty=3,lwd=1)
+lines(1959:2004, fe02,col="red",lty=3,lwd=1)
+
[…]

### 2015-09-21 BPLE/R/LTuniform.R
work done in prague and on train. smoothed, now starting to do more standard calcs

@@ -1,0 +1,55 @@
+
+LTuniform <- function(mx,sex = "f"){
+	#install.packages("/home/tim/git/HMDLifeTables/HMDLifeTables/HMDLifeTables",repos=NULL)
+	require(HMDLifeTables)
+	i.openage <- nrow(mx)
+	ax        <- mx * 0 + .5                                          # ax = .5, pg 38 MPv5
+	
+	ax[1, ]   <- AKm02a0(m0 = mx[1, ], sex = sex)
+	
+#  if (testa0){
+#    ax[1, ]   <- AKm02a0_direct(m0 = mx[1, ], sex = sex)
+#  }
+# multiplying 2 matrices using '*' does the hadamard product in R (elementwise).
+	qx        <- mx / (1 + (1 - ax) * mx)                             # Eq 60 MPv5
+# ---------------------------------------------------------------------------------
+# set open age qx to 1
+	qx[i.openage, ]       <- ifelse(is.na(qx[i.openage, ]), NA, 1)
+	ax[i.openage, ]       <- 1 / mx[i.openage, ]                   
+# ---------------------------------------------------------------------------------
+# define remaining lifetable columns:
+	px 				      <- 1 - qx 																				# Eq 64 MPv5
+	px[is.nan(px)]  <- 0 # skips BEL NAs, as these are distinct from NaNs
+# lx needs to be done columnwise over px, argument 2 refers to the margin.
+	lx 			        <- apply(px, 2, function(px., RADIX, OPENAGE){ 		# Eq 65 MPv5
+				if (all(is.na(px.))) {
+					px.
+				} else {
+					c(RADIX, RADIX * cumprod(px.[1:OPENAGE]))
+				}
+			}, RADIX = 1, OPENAGE = i.openage - 1
+	)
+	rownames(lx)    <- 0:(i.openage - 1) # these got thrown off because l0 imputed.
+# NA should only be possible if there was a death with no Exp below age 80- impossible, but just to be sure
+# lx[is.na(lx)]   <- 0 # removed for BEL testing        
+	dx 				      <- lx * qx 																				# Eq 66 MPv5
+	Lx 				      <- lx - (1 - ax) * dx 														# Eq 67 MPv5
+	
+	Lx[i.openage, ]	<- lx[i.openage, ] * ax[i.openage, ]
[…]

### 2015-09-21 BPLE/R/StatevBPdecomp.R

@@ -1,0 +1,27 @@
+# Author: tim
+###############################################################################
+
+# let's decompose with respect to the e0 that would come from the smoothed mx, mxs
+
+if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
+	# if I'm on the laptop
+	setwd("/home/tim/git/BPLE/BPLE")
+} else {
+	# in that case I'm on Berkeley system, and other people in the dept can run this too
+	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
+}
+library(data.table)
+library(reshape2)
+getwd()
+source("R/LTuniform.R")
+LTC <- local(get(load("Data/LTC1.Rdata")))
+
+.SD <- LTC[LTC$Sex == "f" & LTC$State == "AK" & LTC$Cause == "Cardio.", ]
+table(.SD$Age, .SD$Year)
+gete0 <- function(.SD){
+	mx <- acast(.SD, Age~Year, value.var = "mxs")
+	LTuniform(mx,sex=unique(.SD$Sex))$e0
+}
+### not working so far
+LT <- LTC[LTC$Cause == "Cardio. ",]
+e0sm <- LT[LT,gete0(.SD),by=list(State,Sex)]

### 2015-09-21 BPLE/R/LTuniform.R
code to speed up lifetable calcs, compile continuous decompostion, and create decomposition.

@@ -49,6 +49,25 @@ LTuniform <- function(mx,sex = "f"){
 }
 
 
+LTuniformvecminimal <- compiler::cmpfun(function(mx,sex = "f"){
+	i.openage <- length(mx)
+	OPENAGE   <- i.openage - 1
+	RADIX     <- 1
+	ax        <- mx * 0 + .5
+	ax[1]   <- HMDLifeTables:::AKm02a0(m0 = mx[1], sex = sex)
+	qx        <- mx / (1 + (1 - ax) * mx)
+	qx[i.openage]       <- ifelse(is.na(qx[i.openage]), NA, 1)
+	ax[i.openage]       <- 1 / mx[i.openage]                   
+	px 				    <- 1 - qx
+	px[is.nan(px)]      <- 0
+	lx 			        <- c(RADIX, RADIX * cumprod(px[1:OPENAGE]))
+	dx 				    <- lx * qx
+	Lx 				    <- lx - (1 - ax) * dx
+	Lx[i.openage ]	    <- lx[i.openage ] * ax[i.openage ]
+	Tx 				    <- c(rev(cumsum(rev(Lx[1:OPENAGE]))),0) + Lx[i.openage]
+	ex 				    <- Tx / lx
+	ex[1]
+})
 
 
 

### 2015-09-21 BPLE/R/StatevBPdecomp.R

@@ -18,10 +18,169 @@ LTC <- local(get(load("Data/LTC1.Rdata")))
 
 .SD <- LTC[LTC$Sex == "f" & LTC$State == "AK" & LTC$Cause == "Cardio.", ]
 table(.SD$Age, .SD$Year)
-gete0 <- function(.SD){
-	mx <- acast(.SD, Age~Year, value.var = "mxs")
-	LTuniform(mx,sex=unique(.SD$Sex))$e0
+Age <- .SD$Age
+Year <- .SD$Year
+mx <- .SD$mxs
+Sex <- .SD$Sex
+
+gete0 <- function(CHUNK){
+	mx <- acast(CHUNK, Age~Year, value.var = "mxs")
+	data.frame(State = unique(CHUNK$State),
+			   Year = sort(unique(CHUNK$Year)),
+			   Sex = unique(CHUNK$Sex),
+			   e0 = LTuniform(mx,sex=unique(CHUNK$Sex))$e0,
+			   e0orig = CHUNK$ex[CHUNK$Age ==0],
+			   stringsAsFactors = FALSE)
 }
-### not working so far
-LT <- LTC[LTC$Cause == "Cardio. ",]
-e0sm <- LT[LT,gete0(.SD),by=list(State,Sex)]
+gete02 <- function(.SD){
+	mx <- acast(.SD, Age~Year, value.var = "mxs")
+	c(LTuniform(mx,sex=unique(.SD$Sex))$ex)
+}
+
+# for some reason couldn't do this with
+# normal data.table. tried too much, so I gave up
+LT       <- LTC[LTC$Cause == "Cardio.",]
+LT       <-as.data.frame(LT)
+LTL      <- split(LT, list(LT$State,LT$Sex))
+e0states <- do.call(rbind, lapply(LTL, gete0))
+head(e0states)
+e0ms <- acast(e0states[e0states$Sex == "m", ],State~Year,value.var = "e0")
+e0m  <- acast(e0states[e0states$Sex == "m", ],State~Year,value.var = "e0orig")
+e0fs <- acast(e0states[e0states$Sex == "f", ],State~Year,value.var = "e0")
+e0f  <- acast(e0states[e0states$Sex == "f", ],State~Year,value.var = "e0orig")
+
+matplot(1959:2004, t(e0m), type = 'l', lty = 2, lwd=2, col = "#00000090")
[…]

### 2015-09-21 BPLE/R/lookatdata.R

@@ -31,7 +31,7 @@ States <- data.table(States)
 Mins   <- States[,min(mxcs2),by=list(Year, Sex, Age, Cause)]
 Mins2  <- States[,min(mxc5),by=list(Year, Sex, Age, Cause)]
 
-Agg <- Mins[,sum(V1),by=list(Year, Sex, Age)]
+Agg  <- Mins[,sum(V1),by=list(Year, Sex, Age)]
 Agg2 <- Mins2[,sum(V1),by=list(Year, Sex, Age)]
 
 library(reshape2)

### 2015-09-21 BPLE/R/mydecomp.R

@@ -1,0 +1,5 @@
+
+# Author: tim
+###############################################################################
+
+mydecomp <- compiler::cmpfun(DecompHoriuchi::DecompContinuousOrig)

### 2015-09-22 BPLE/R/CauseStateDecade.R
plot state year departures from BP by decades and causes

@@ -1,0 +1,59 @@
+
+# Author: tim
+###############################################################################
+
+
+# find total contribution from Cause C, State S by decade.
+
+if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
+	# if I'm on the laptop
+	setwd("/home/tim/git/BPLE/BPLE")
+} else {
+	# in that case I'm on Berkeley system, and other people in the dept can run this too
+	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
+}
+
+fem <- local(get(load("Data/fcontrib.Rdata")))
+mal <- local(get(load("Data/mcontrib.Rdata")))
+
+names(fem)
+names(fem[[1]])
+names(fem)
+ftot <- lapply(names(fem),function(nm,fem){
+			do.call(rbind,lapply(fem,colSums))
+		}, fem = fem)
+
+sts     <- names(fem[[1]])
+decades <- (1960:2004) - (1960:2004) %% 10
+decs    <- data.frame(Year = 1960:2004,decade = decades)
+decs$Year
+StateCauseYear <- do.call(rbind, lapply(sts, function(st,decs){
+					decs$State <- st
+					decs
+				}, decs=decs))
+MSCY <- FSCY <- StateCauseYear
+
+Causes <- colnames(fem[[1]][[1]])
+Causes <- matrix(0,ncol=length(Causes),nrow=nrow(StateCauseYear), dimnames=list(NULL, Causes))
+MSCY <- cbind(MSCY, Causes)
+FSCY <- cbind(FSCY, Causes)
+MSCY$Uterus <- NULL
+
+for (i in 1:nrow(MSCY)){
+	yr   <- as.character(MSCY$Year[i])
+	St   <- MSCY$State[i]
+	MSums <- colSums(mal[[yr]][[St]])
+	FSums <- colSums(fem[[yr]][[St]])
+	MSCY[i,names(MSums)]  <- MSums
+	FSCY[i,names(FSums)] <- FSums
+}
+
+save(MSCY, file = "Data/MSCY.Rdata")
+save(FSCY, file = "Data/FSCY.Rdata")
+
+
+
+
+
+
+

### 2015-09-22 BPLE/R/StatesHeat.R

@@ -1,0 +1,111 @@
+# Author: tim states square heatmaps.
+###############################################################################
+source("R/colourtools.R")
+
+# custom, for artisanal panel plots
+centercoords <- function(x1,x2,y1,y2){
+	xrange <- range(c(x1,x2))
+	yrange <- range(c(y1,y2))
+	div    <- max(c(xrange,yrange))
+	x1     <- x1/div - .5
+	x2     <- x2/div - .5
+	y1     <- y1/div - .5
+	y2     <- y2/div - .5
+	
+	list(x1=x1,x2=x2,y1=y1,y2=y2)
+}
+'%==%' <- function(x,y){
+	!is.na(x) & !is.na(y) & x == y
+}
+#ST <- matrix(c("AK",rep("",10),"ME",
+#		rep("",10),"VT","NH",
+#		"","WA","ID","MT","ND","MN","IL","WI","MI","NY","RI","MS",
+#		"","OR","NV","WY","SD","IA","IN","OH","PA","NJ","CT","",
+#		"","CA","UT","CO","NB","MO","KY","WV","VA","MD","DL","",
+#		"","","AZ","NM","KS","AR","TN","NC","SC","DC","","",
+#		"","","","","OK","LA","MS","AL","GA","","","",
+#		"HA","","","","TX","","","","","FL","",""),ncol = 12, nrow = 8,byrow=TRUE)
+#
+#ST[ST == ""] <- NA
+#Color <- matrix(NA,nrow=nrow(ST),ncol=ncol(ST))
+#Color[!is.na(ST[8:1,])] <- gray(.5)
+#
+#x1 <- col(ST) - 1
+#x2 <- col(ST)
+#y1 <- row(ST) - 1
+#y2 <- row(ST)
+#
+#coords <- centercoords(x1,x2,y1,y2)
+#
+#
+#plot(NULL, xlim = c(-.5,.5),ylim=c(-.5,.5),xlab = "", ylab = "", type= "n", axes = FALSE,asp=1)
+#with(coords,rect(x1,y1,x2,y2,border="white",col = Color))
+#with(coords,text((x1+x2)/2,(y1+y2)/2,ST[8:1,],col="white"))
+
+#install.packages("spatstats")
+#require(devtools)
+#install_github('spatstat/spatstat')
+# use to determine color for printing state name
[…]

### 2015-09-22 BPLE/R/StatesSquares.R

@@ -1,29 +1,134 @@
 
-# Author: tim
-###############################################################################
+if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
+	# if I'm on the laptop
+	setwd("/home/tim/git/BPLE/BPLE")
+} else {
+	# in that case I'm on Berkeley system, and other people in the dept can run this too
+	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
+}
+source("R/StatesHeat.R")
+
+FSCY <- local(get(load("Data/FSCY.Rdata")))
+MSCY <- local(get(load("Data/MSCY.Rdata")))
+
+FSCY <- data.table(FSCY)
+MSCY <- data.table(MSCY)
+
+setnames(FSCY,colnames(FSCY),c("Year","Decade","State","All_other","Breast","Cardio","Cerebrov",
+				"Lung","Other_CVDs","Other_MN","Other_smoking","Prostate","Stomach","Uterus"))
+setnames(MSCY,colnames(MSCY),c("Year","Decade","State","All_other","Breast","Cardio","Cerebrov",
+				"Lung","Other_CVDs","Other_MN","Other_smoking","Prostate","Stomach"))
+
+# aggregate: take decade means
+FSCY <- FSCY[, list(All_other=mean(All_other), 
+				   Breast=mean(Breast),
+				   Cardio=mean(Cardio),
+				   Cerebrov=mean(Cerebrov),
+				   Lung=mean(Lung),
+				   Other_CVDs=mean(Other_CVDs),
+				   Other_MN=mean(Other_MN),
+				   Other_smoking=mean(Other_smoking),
+				   Stomach=mean(Stomach),
+				   Uterus=mean(Uterus)), 
+		by=list(State, Decade)]
+Causesf <- colnames(FSCY)[-c(1:2)]
+MSCY <- MSCY[, list(All_other=mean(All_other), 
+				Breast=mean(Breast),
+				Cardio=mean(Cardio),
+				Cerebrov=mean(Cerebrov),
+				Lung=mean(Lung),
[…]

### 2015-09-22 BPLE/Manuscript.tex
add references, manuscript, and some figures

@@
-% an empty .tex file. Eventually may transwer .odt in here
+% an empty .tex file. Eventually may transfer .odt in here
+%%This is a very basic article template.
+%%There is just one section and two subsections.
+%\documentclass[12pt,oneside,a4paper,doublespacing]{article} % for submission
+\documentclass[11pt,oneside,a4paper]{article} % for sharing
+
+\usepackage{appendix}
+\usepackage{amsmath}
+\usepackage{caption}
+\usepackage{placeins}
+\usepackage{graphicx}
+\usepackage{subcaption}
+%\usepackage{subfig}
+\usepackage{longtable}
+\usepackage{setspace}
+%\usepackage{tikz}
+\usepackage{booktabs}
+\usepackage{tabularx}
+\usepackage{xcolor,colortbl}
+\usepackage{chngpage}
+%\usepackage[active,tightpage]{preview}
+\usepackage{natbib}
+\bibpunct{(}{)}{,}{a}{}{;} 
+\usepackage{url}
+\usepackage{nth}
+\usepackage{authblk}
+\usepackage[most]{tcolorbox}
+%\usepackage{hyperref}
+%\usepackage{color}
+%\usepackage{fontspec}
+%\usepackage{pdfsync}
+\usepackage[normalem]{ulem}
+\usepackage{amsfonts}
+\renewcommand{\listtablename}{List of Appendix Tables}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+% working on this need to concatenate file name based on sex and variable name
+%\newcommand\Cell[1]{{\raisebox{-0.05in}{\includegraphics[height=.2in,width=.2in]{Figures/ColorCodes/\expandafter#1}}}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% setting color to letters affects spacing. Here's a hack I found here:
+% http://tex.stackexchange.com/questions/212736/change-letter-colour-without-losing-letter-spacing
+%\DeclareRobustCommand{\spacedallcaps}[1]{\MakeUppercase{\textsc{#1}}} % all
+% caps with better spacing
+
+%\colorlet{RED}{red}
+%\colorlet{BLUE}{b}
+\colorlet{rd}{red}
+\colorlet{bl}{blue}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand\ackn[1]{%
+  \begingroup
+  \renewcommand\thefootnote{}\footnote{#1}%
+  \addtocounter{footnote}{-1}%
+  \endgroup
+}
+\newcommand\vt[1]{\textcolor{rd}{#1}}
+\newcommand\eg[1]{\textcolor{bl}{#1}}
+
+\newcommand\tg[1]{\includegraphics[scale=.5]{Figures/triadtable/triad#1.pdf}}
+\newcommand\tgh[1]{\raisebox{-.25\height}{\includegraphics[scale=.3]{Figures/triadtable/triad#1.pdf}}}
+
+\defcitealias{HMD}{HMD}
+
+% junk for longtable caption
+\AtBeginEnvironment{longtable}{\linespread{1}\selectfont}
+\setlength{\LTcapwidth}{\linewidth}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
+
+\title{A best practices composite lifetable for US states}
+
+\author[1]{Tim Riffe\thanks{triffe@demog.berkeley.edu}}
+\author[2,3]{Adrien Remund}
+\author[2,3]{Magali Barbieri}
+\author[3]{Celeste Winant}
+\affil[1]{Max Planck Institute for Demographic Research}
+\affil[2]{universit{\'e} de Gen{\`e}ve}
+\affil[2]{Institut National d'{\'e}tudes D{\'e}mgraphiques}
+\affil[3]{Department of Demography, University of California, Berkeley}
+
+%\author{[Authors]}
+
+\maketitle
+
+\begin{abstract}
+We calculate a preliminary series of best-possible lifetables for the United
+States from 1959-2004, defined as the age-specific aggregate of the lowest
+observed age-cause specific death rates among the 50
+states and the District of Columbia for each year. This synthetic best practices lifetable shows a gradual
+increasing trend over the period, on average 1.9 and 2.2 years higher
+than the highest state life expectancy in each year for males and females,
+respectively. We argue that the US states best practices lifetable is a useful
+guage of mortality.
+\ackn{The work reported in this manuscript was supported in part by the U.S.
+National Institute On Aging of the National Institutes of Health under award
+numbers R01-AG011552 and R01-AG040245. The content is solely the responsibility of the authors and does not necessarily represent the official views of the funding agencies.}
+\end{abstract}
+
+\section*{Introduction}
+
+The question of limits to life expectancy is fundamental to demography, but it
+is also practical when projecting mortality. Mortality reductions in past
+decades have been steady in many populations, and at times linear, which tends
+to guide projections into predicting the same sort of progress far into the
+future. Most past attempts to place limits on such projections were shown to be
+overly conservative within a short period of years \citep{oeppen2002broken}.
+However, many simple mortality projections will send life expectancy scenarios
+to very high levels that for a given population may seem unimaginable. In some
+cases, one seeks an external population that has already achieved a life
+expectancy as high as that produced by a model, and this gives some assurance
+that the projection is possible. Japan assumes this role in many cases today. In
+2012 life expectancy for Japanese females was 86, while for US females it was 81
+\citep{HMD}. If a US projection turns out to reach 86 we at least know that this
+is indeed possible.
+
+The desire for such external references explains part of the interest in the
+historical development of world record life expectancy \citep{oeppen2002broken}.
+The maximum life expectancy from a set of populations shows what others may one
+day achieve due to diffusion in practices, technology, and wellbeing
+\citep{vallin2010esperance}.
+The vanguard life expectancy therefore provides a benchmark. The vanguard life expectancy is calculated based on mortality rates undifferentiated
+by cause, which may not provide an omnibus signal of what may eventually develop.
+While the force of mortality governs the lifetable, there is a substantive rationale to conceive of this force as a composite of cause-specific forces of mortality.
+\citet{vallin2008minimum} differentiate trends in life expectancy by causes of death, in terms of an epidemiological transition that unfolds in progressive stages with respect to particular technologies and risk factors in recent history.
+The level and timing of cause-specific responses to particular technological or well-being improvements at times vary.
+
+The lowest all-cause force of mortality is not necessarily composed of the
+lowest cause-specific forces of mortality, which means that the vanguard life
+expectancy does not reflect the best mortality possible given the current state
+of conditions. Alternatively, one could create cause-specific vanguard
+benchmarks, or more synthetically, combine the lowest mortality observed by age
+and cause into a hypothetical minimum mortality schedule. Such a hybrid
+lifetable-- hybrid with respect to reference population-- was already suggested
+by \citet{wunsch1975minimum} and \citet{vallin2008minimum}. These authors
+investigated trends in national-level. We refer to such a hybrid minimum
+lifetable as a best practices lifetable.
+
+National populations are heterogeneous with respect to many factors that affect cause-spectific mortality rates.
+This does not make such comparisons futile, but it may make perfect convergence unimaginable even in the very long run.
+A national population's best reference may be calculated from within its own territorial subpopulations.
+This practice limits the sources of unaccounted for heterogeneity in international comparisons.
+However, subnational territorial units dto are not necessarily statistically reliable for this sort of exercise due to stochastic fluctuations in small populations, and varying degrees of data quality.
+The United States has a large population, with 50 states plus the District of Columbia that are also large enough to reduce stochasticity to male cause-specific mortality usable.
+Further the United States has a homogeneous death registration system with  standardized cause coding classification and practice.
+These conveniences with respect to the data source avoid many of the pitfalls of international comparisons of cause of death data.
+We therefore compute the US best practice lifetable based on 51 territorial subpopulations and 12 causes of death for the period 1959 to 2004 (or 2013 if RDC cooperates)
+
+\end{document}

### 2015-09-22 BPLE/R/StatesSquares.R

@@ -31,7 +31,6 @@ FSCY <- FSCY[, list(All_other=mean(All_other),
 				   Stomach=mean(Stomach),
 				   Uterus=mean(Uterus)), 
 		by=list(State, Decade)]
-Causesf <- colnames(FSCY)[-c(1:2)]
 MSCY <- MSCY[, list(All_other=mean(All_other), 
 				Breast=mean(Breast),
 				Cardio=mean(Cardio),
@@ -43,10 +42,25 @@ MSCY <- MSCY[, list(All_other=mean(All_other),
 				Prostate=mean(Prostate),
 				Stomach=mean(Stomach)),
 				by=list(State, Decade)]
-Causesm <- colnames(MSCY)[-c(1:2)]
+
 
 MSCY   <- as.data.frame(MSCY)
 FSCY   <- as.data.frame(FSCY)
+
+MSCY$Lung          <- MSCY$Lung + MSCY$Other_smoking
+FSCY$Lung          <- FSCY$Lung + FSCY$Other_smoking
+MSCY$Other_smoking <- NULL
+FSCY$Other_smoking <- NULL
+
+FSCY$Cancer 	<- FSCY$Breast + FSCY$Other_MN + FSCY$Stomach + FSCY$Uterus
+FSCY$Breast 	<- FSCY$Other_MN <- FSCY$Stomach <- FSCY$Uterus <- NULL
+FSCY$Cerebrov 	<- FSCY$Cerebrov + FSCY$Other_CVDs
+FSCY$Other_CVDs <- NULL
+
+MSCY$Cancer   	<- MSCY$Other_MN + MSCY$Prostate + MSCY$Stomach + MSCY$Breast
+MSCY$Other_MN 	<- MSCY$Prostate <- MSCY$Stomach <- MSCY$Breast <- NULL
+MSCY$Cerebrov 	<- MSCY$Cerebrov + MSCY$Other_CVDs
+MSCY$Other_CVDs <- NULL
 
 # for purposes of plotting, top off upper contrib
 MSCY[,-c(1,2)][MSCY[,-c(1,2)] > 2] <- 2
@@ -70,6 +84,9 @@ F1980s$State <- F1980s$Decade <- NULL
 F1990s$State <- F1990s$Decade <- NULL
 F2000s$State <- F2000s$Decade <- NULL
 
+#colSums(MSCY[,-c(1,2)])
+#colSums(FSCY[,-c(1,2)])
+
 M1960s <- MSCY[MSCY$Decade == 1960, ]
[…]

### 2015-09-22 BPLE/R/StatevBPdecomp.R

@@ -23,37 +23,37 @@ Year <- .SD$Year
 mx <- .SD$mxs
 Sex <- .SD$Sex
 
-gete0 <- function(CHUNK){
-	mx <- acast(CHUNK, Age~Year, value.var = "mxs")
-	data.frame(State = unique(CHUNK$State),
-			   Year = sort(unique(CHUNK$Year)),
-			   Sex = unique(CHUNK$Sex),
-			   e0 = LTuniform(mx,sex=unique(CHUNK$Sex))$e0,
-			   e0orig = CHUNK$ex[CHUNK$Age ==0],
-			   stringsAsFactors = FALSE)
-}
-gete02 <- function(.SD){
-	mx <- acast(.SD, Age~Year, value.var = "mxs")
-	c(LTuniform(mx,sex=unique(.SD$Sex))$ex)
-}
-
-# for some reason couldn't do this with
-# normal data.table. tried too much, so I gave up
-LT       <- LTC[LTC$Cause == "Cardio.",]
-LT       <-as.data.frame(LT)
-LTL      <- split(LT, list(LT$State,LT$Sex))
-e0states <- do.call(rbind, lapply(LTL, gete0))
-head(e0states)
-e0ms <- acast(e0states[e0states$Sex == "m", ],State~Year,value.var = "e0")
-e0m  <- acast(e0states[e0states$Sex == "m", ],State~Year,value.var = "e0orig")
-e0fs <- acast(e0states[e0states$Sex == "f", ],State~Year,value.var = "e0")
-e0f  <- acast(e0states[e0states$Sex == "f", ],State~Year,value.var = "e0orig")
-
-matplot(1959:2004, t(e0m), type = 'l', lty = 2, lwd=2, col = "#00000090")
-matplot(1959:2004, t(e0ms), type = 'l', lty =1, col = "#FF000050", add=TRUE)
-
-matplot(1959:2004, t(e0f), type = 'l', lty = 2, lwd=2, col = "#00000090")
-matplot(1959:2004, t(e0fs), type = 'l', lty =1, col = "#FF000050", add=TRUE)
+#gete0 <- function(CHUNK){
+#	mx <- acast(CHUNK, Age~Year, value.var = "mxs")
+#	data.frame(State = unique(CHUNK$State),
+#			   Year = sort(unique(CHUNK$Year)),
+#			   Sex = unique(CHUNK$Sex),
+#			   e0 = LTuniform(mx,sex=unique(CHUNK$Sex))$e0,
[…]

### 2015-09-22 BPLE/R/e0trends.R

@@ -1,0 +1,93 @@
+if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
+	# if I'm on the laptop
+	setwd("/home/tim/git/BPLE/BPLE")
+} else {
+	# in that case I'm on Berkeley system, and other people in the dept can run this too
+	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
+}
+getwd()
+
+LTC <- local(get(load("Data/LTC1.Rdata")))
+
+gete0 <- function(CHUNK){
+	mx <- acast(CHUNK, Age~Year, value.var = "mxs")
+	data.frame(State = unique(CHUNK$State),
+			Year = sort(unique(CHUNK$Year)),
+			Sex = unique(CHUNK$Sex),
+			e0 = LTuniform(mx,sex=unique(CHUNK$Sex))$e0,
+			e0orig = CHUNK$ex[CHUNK$Age ==0],
+			stringsAsFactors = FALSE)
+}
+gete03 <- function(mx,Sex){
+	LTuniformvecminimal(mx,sex=unique(Sex))
+}
+
+# for some reason couldn't do this with
+# normal data.table. tried too much, so I gave up
+States <- LTC[,sum(mxcs2),by=list(State,Year,Sex,Age)]
+setnames(States,"V1","mxs")
+
+e0states <- States[,gete03(mxs,Sex), by = list(State,Year,Sex)]
+setnames(e0states,"V1","e0")
+
+# now get BPLE
+Mins <- LTC[,min(mxcs2),by=list(Year, Sex, Age, Cause)]
+BPmx  <- Mins[,sum(V1),by=list(Year, Sex, Age)]
+setnames(BPmx,"V1","mxmin")
+e0bp  <- BPmx[, gete03(mxmin,Sex), by = list(Year,Sex)]
+setnames(e0bp,"V1","e0")
+
+
+e0ms <- acast(e0states[e0states$Sex == "m", ],State~Year,value.var = "e0")
+bpm  <- e0bp$e0[e0bp$Sex == "m"]
+e0fs <- acast(e0states[e0states$Sex == "f", ],State~Year,value.var = "e0")
+bpf  <- e0bp$e0[e0bp$Sex == "f"]
+
+# now US aggregate.
[…]

### 2015-09-22 BPLE/R/lookatdata.R

@@ -18,7 +18,11 @@ if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
 getwd()
 
 States <- local(get(load("Data/LTC1.Rdata")))
-LT     <- local(get(load("Data/MilaStatesSingleAge.Rdata")))
+#LT     <- local(get(load("Data/MilaStatesSingleAge.Rdata")))
+
+
+fRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "f"], LT$Year[LT$Age == 0& LT$Sex == "f"], max)
+mRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "m"], LT$Year[LT$Age == 0& LT$Sex == "m"], max)
 
 fRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "f"], LT$Year[LT$Age == 0& LT$Sex == "f"], max)
 mRLE <- tapply(LT$ex[LT$Age == 0 & LT$Sex == "m"], LT$Year[LT$Age == 0& LT$Sex == "m"], max)

### 2015-09-22 BPLE/build.R
include junky pdf for now

@@ -1,0 +1,5 @@
+
+#setwd("/home/tim/git/BPLE/BPLE")
+system("pdflatex Manuscript.tex Manuscript.pdf")
+
+

### 2015-09-23 BPLE/Manuscript.tex
include junky pdf for now
updating abstract

@@
 \author[2,3]{Adrien Remund}
-\author[2,3]{Magali Barbieri}
-\author[3]{Celeste Winant}
+\author[3,4]{Magali Barbieri}
+\author[4]{Celeste Winant}
 \affil[1]{Max Planck Institute for Demographic Research}
 \affil[2]{universit{\'e} de Gen{\`e}ve}
-\affil[2]{Institut National d'{\'e}tudes D{\'e}mgraphiques}
-\affil[3]{Department of Demography, University of California, Berkeley}
+\affil[3]{Institut National d'{\'e}tudes D{\'e}mgraphiques}
+\affil[4]{Department of Demography, University of California, Berkeley}
 
@@
 historical development of world record life expectancy \citep{oeppen2002broken}.
-The maximum life expectancy from a set of populations shows what others may one
-day achieve due to diffusion in practices, technology, and wellbeing
-\citep{vallin2010esperance}.
-The vanguard life expectancy therefore provides a benchmark. The vanguard life expectancy is calculated based on mortality rates undifferentiated
-by cause, which may not provide an omnibus signal of what may eventually develop.
-While the force of mortality governs the lifetable, there is a substantive rationale to conceive of this force as a composite of cause-specific forces of mortality.
-\citet{vallin2008minimum} differentiate trends in life expectancy by causes of death, in terms of an epidemiological transition that unfolds in progressive stages with respect to particular technologies and risk factors in recent history.
-The level and timing of cause-specific responses to particular technological or well-being improvements at times vary.
+The maximum life expectancy from a set of populations shows what other
+populations may one day achieve due to diffusion in practices, technology, and
+wellbeing \citep{vallin2010esperance}. The vanguard life expectancy is
+calculated based on mortality rates undifferentiated by cause of death, which
+may not provide an omnibus signal of what may eventually develop. While the force of
+mortality governs the lifetable, there is a substantive rationale to conceive of
+this force as a composite of cause-specific forces of mortality.
+\citet{vallin2008minimum} separate trends in life expectancy by causes of death,
+explained in terms of an epidemiological transition that unfolds in progressive
+stages with respect to particular technologies and risk factors in recent
+history.
 
-The lowest all-cause force of mortality is not necessarily composed of the
-lowest cause-specific forces of mortality, which means that the vanguard life
-expectancy does not reflect the best mortality possible given the current state
-of conditions. Alternatively, one could create cause-specific vanguard
-benchmarks, or more synthetically, combine the lowest mortality observed by age
-and cause into a hypothetical minimum mortality schedule. Such a hybrid
-lifetable-- hybrid with respect to reference population-- was already suggested
-by \citet{wunsch1975minimum} and \citet{vallin2008minimum}. These authors
-investigated trends in national-level. We refer to such a hybrid minimum
-lifetable as a best practices lifetable.
+The level and timing of cause-specific responses to particular
+technological or well-being improvements varies between populations. Therefore,
+the lowest observed all-cause force of mortality is not necessarily composed of
+the lowest cause-specific forces of mortality, which means that the vanguard life
+expectancy may not reflect the best mortality possible given current conditions
+over a set of populations. For instance, the lowest rate of suicide and heart
+disease at age 45 will not necessarily be found in the same population. As an
+alternative to vanguard life expectancy, one could create cause-specific
+low-mortality benchmarks, or more synthetically, combine the lowest mortality
+observed by age and cause into a hybrid minimum mortality schedule.
+This synthetic low mortality lifetable--- hybrid with respect to reference population over age and cause--- was already suggested by \citet{wunsch1975minimum} and \citet{vallin2008minimum}.
+These authors investigated trends at the national-level.
+We refer to such hybrid minimum
+lifetables as best practices lifetables.
 
-National populations are heterogeneous with respect to many factors that affect cause-spectific mortality rates.
-This does not make such comparisons futile, but it may make perfect convergence unimaginable even in the very long run.
-A national population's best reference may be calculated from within its own territorial subpopulations.
-This practice limits the sources of unaccounted for heterogeneity in international comparisons.
-However, subnational territorial units dto are not necessarily statistically reliable for this sort of exercise due to stochastic fluctuations in small populations, and varying degrees of data quality.
-The United States has a large population, with 50 states plus the District of Columbia that are also large enough to reduce stochasticity to male cause-specific mortality usable.
-Further the United States has a homogeneous death registration system with  standardized cause coding classification and practice.
-These conveniences with respect to the data source avoid many of the pitfalls of international comparisons of cause of death data.
-We therefore compute the US best practice lifetable based on 51 territorial subpopulations and 12 causes of death for the period 1959 to 2004 (or 2013 if RDC cooperates)
+National populations are heterogeneous with respect to many factors, both
+observed and unobserved, that affect cause-spectific mortality rates. This does
+not make international mortality comparisons futile, but it may make perfect
+convergence unimaginable even in the very long run. This is the case for both
+all cause and cause-specific mortality. When projecting mortality for a given country, we wonder whether mortality from an external population, or set of populations, is the best ``sanity
+check'' for mortality forecasts.
+For this reason, we propose a minimum mortality trend for the United States
+calculated on the basis of its own territorial subdivisions, the 50 states plus
+the District of Columbia. This choice limits the sources of unaccounted for
+heterogeneity in international comparisons, and it also conveniently homogenizes
+data sources. To be clear, we do not necessarily propose using a trend in best
+practices mortality as a component of coherent mortality forecasts, and we do
+not undertake such an exercise. Instead, we view the U.S. best practices
+mortality as indicative of the potential already present in the U.S., a special
+kind of ``nowcast''.
 
+\section*{Data and methods}
+All death count data come from the public NCHS mortality microdata files
+\citep{NCHSdata}. Population denominators come from a beta version of the U.S.
+States project of the Human Mortality Database \citep{HMD}. Further processing of
+population counts into exposures is done according to the HMD Methods
+Protocol, and all-cause lifetables were calculated by single ages \citep{HMDMP}.
+Death counts for causes were aggregated into quinquennial age groups by year and sex for eleven large groups that minimize disruptions between ICD transitions, and then
+converted to fractions. Fractions were then multiplied into the single-age
+all-cause death rates.
+
+One caveat of calculating a best practices lifetable is that stochastic zeros
+occur in particular ages and causes, especially among small populations. The
+exercise of selecting the minimum death rate for a given age and cause will tend
+to collect stochastic zeros, producing an excessively optimistic best practices
+life expectancy. We therefore smooth all death rates using
+the \texttt{MortalitySmooth} package \citep{GC2012} in \texttt{R}. The age-period
+matrix of death rates for each state and sex is smoothed over both margins,
+fractions are then recalculated, a multiplied into the smoothed all-cause death
+rates. Finally, we calculate lifetables for both states and best practices
+lifetables using HMD methodology.
+
+We decompose differences in state life expectancy from the best-practices life
+expectancy using the pseudo-continuous method proposed by
+\citet{horiuchi2008decomposition}. We prefer this method because it allows for
+defining life expectancy as a direct function of age-cause-specific rates,
+$M_{(x,c)}$, thereby eliminating some artifacts present in other methods.
+Results are summarized by ages, causes and states and can be interpreted
+directly as the the total lag in years of life expectancy due to excess
+mortality (anything greater than the minimum) in the given margin.\footnote{In
+the final version we may opt to further separate trends from initial
+differences using the decomposition method recently proposed by
+\citet{DimaDecomp2014}.}
+
+\FloatBarrier
+\section*{Preliminary results}
+\FloatBarrier
+
+All results at this time are preliminary, but we can report broad trends at this
+time.\footnote{The final results will differ in the following ways. At this time we have
+calculated results using 11 causes of death that were previously used in a different research project, and this will
+change to 30 causes to be consistent and comparable with \citet{vallin2008minimum}. Including
+more causes will increase the best practices life expectancy level. We will also
+include estimates through 2012 or 2013. At this time, we were only able to
+produce results through 2004, because geographic information is suppressed in
+the microdata for later years. This information is available in secure Research
+Data Centers, where we will produce cause fractions and all-cause lifetables for
+the most recent decade.}
+
+\subsection*{Life expectancy trends}
+Most states display irregular gradual increasing trends
+over the 45 years considered, except for males in the District of Columbia,
+which has been recovering strongly since the 1990s, due both to secular
+improvements and compositional change. The US aggregate trend is bouncier
+because death counts were not smoothed, and its rank position among states
+varies due in part also to compositional changes between states. The vanguard
+state life expectancy for males is Hawaii until the two most recent years of
+observation (2003-2004), where Minnesota has taken the lead. For females, the
+Minnesota is vanguard until 1963, followed by North Dakota for five years, and
+then Hawaii for the rest of the series until 2004. The vanguard trend for
+females has the same overall shape as the cluster of individual state trends,
+but for males the time trend in vanguard, ergo Hawaiian, life expectancy was
+different from the bulk of states.
+
+For both males and females, the best
+practices trend is on average 2 years higher than the vanguard trend, and its
+progress is steadier than either the vanguard or national trends, as well as
+most individual states. The best practices trend has increased each year for
+males and females, though faster for males. The average annual improvement for
+males has been 0.1790 years of $e(0)$ per annum, compared with 0.1785 for the
+male national average life expectancy. The respective rates of improvement for
+females have been 0.1470 and 0.1456.
+
+\begin{figure}[t!]
+\caption{Trends in US, vanguard, best practices, and state life expectancies
+($e(0)$), 1959-2004.}
+\label{fig:e0}
+\centering
+\begin{subfigure}[b]{.48\textwidth}
+\centering
+\caption{Males}
+\label{fig:e0m}
+\includegraphics[scale=0.5]{Figures/e0trendsM.pdf}
+\end{subfigure}
+~
+\begin{subfigure}[b]{.48\textwidth}
+\centering
+\caption{Females}
+\label{fig:e0f}
+\includegraphics[scale=0.5]{Figures/e0trendsF.pdf}
+\end{subfigure}
+\end{figure}
+
+\subsection{State differences by cause}
+
+We decompose the state departures in life expectancy from the best practice
+trend in each year and sex, over both age and cause. In Figure~\ref{fig:dep}
+we aggregate over age and average within decades to summarize the total
+contribution to the departure from best practices.\footnote{The 2000s trend is
+based on the average for the years 2000-2004 only.} Major causes of death are in
+rows, and decades in columns.
+Part of the attraction of decomposing with respect to the best practices life expectancy is that contributions are always
+of the same sign, which makes interpretation of the heatmap more intuitive.
+Hotter colors indicate larger departures. For the case of cardiovascular disease
+and other causes, grouping departures into the 2$+$ group blends out much
+interstate variation, since some contributions range as high as 4
+years.\footnote{This visual display may be significantly adapted when we add
+more causes to the decomposition. We will also do a similar small multiples
+display for grouped age patterns.}
+
+Since the cause groupings available at the time of submission are not the final groupings, and
+because the causes included in the study contribute such varying magnitudes to
+state differences, trends are dominated by Cardivascular diseases and the large
+aggregate group of all other causes. The 11 original causes were grouped so as
+to make state patterns more visible.
+
+\begin{figure}[t!]
+\caption{State departures from vanguard $e(0)$ by large cause groups,
+1959-2004.}
+\label{fig:dep}
+\centering
+\begin{subfigure}[b]{\textwidth}
+\centering
+\caption{Males}
+\label{fig:depm}
+\includegraphics[scale=0.5]{Figures/StatesDecadesM.pdf}
+\end{subfigure}
+\\
+\begin{subfigure}[b]{\textwidth}
+\centering
+\caption{Females}
+\label{fig:depf}
+\includegraphics[scale=0.5]{Figures/StatesDecadesF.pdf}
+\end{subfigure}
+\end{figure}
+
+\FloatBarrier
+
+\bibliographystyle{plainnat}
+%\bibliographystyle{demography}
+  \bibliography{references.bib}   % Use the BibTeX file ``references.bib''.
 \end{document}

### 2015-09-23 BPLE/R/e0trends.R

@@ -6,18 +6,12 @@ if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
 	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
 }
 getwd()
+library(data.table)
+source("/home/tim/git/BPLE/BPLE/R/LTuniform.R")
 
 LTC <- local(get(load("Data/LTC1.Rdata")))
 
-gete0 <- function(CHUNK){
-	mx <- acast(CHUNK, Age~Year, value.var = "mxs")
-	data.frame(State = unique(CHUNK$State),
-			Year = sort(unique(CHUNK$Year)),
-			Sex = unique(CHUNK$Sex),
-			e0 = LTuniform(mx,sex=unique(CHUNK$Sex))$e0,
-			e0orig = CHUNK$ex[CHUNK$Age ==0],
-			stringsAsFactors = FALSE)
-}
+
 gete03 <- function(mx,Sex){
 	LTuniformvecminimal(mx,sex=unique(Sex))
 }
@@ -90,4 +84,13 @@ max(apply(e0ms,1,function(x){
 mean(bpm - apply(e0ms,2,max))
 mean(bpf - apply(e0fs,2,max))
 
+rownames(e0ms)[apply(e0ms,2,which.max)]
+# HA until MN took over in 2003
+rownames(e0fs)[apply(e0fs,2,which.max)]
+# MN until 53, ND for 5 yrs, then HI
+diff(range(bpm))/45;diff(c(use0m[1],use0m[46]))/46
+diff(range(bpf))/45;diff(c(use0f[1],use0f[46]))/46
+
+diff(c(use0m[1],use0m[46]))/46
+diff(c(use0f[1],use0f[46]))/46
 

### 2015-09-23 BPLE/build.R

@@ -1,5 +1,8 @@
 
 #setwd("/home/tim/git/BPLE/BPLE")
 system("pdflatex Manuscript.tex Manuscript.pdf")
+system("bibtex Manuscript.aux") # to generate .bbl
+system("pdflatex Manuscript.tex Manuscript.pdf")
+system("pdflatex Manuscript.tex Manuscript.pdf")
 
-
+file.copy("Manuscript.pdf","/home/tim/Dropbox/BPLE/Manuscript/Manuscript.pdf",overwrite=TRUE)

### 2015-09-23 BPLE/Manuscript.tex
final commit before sending to magali. fol din changes tomorrow

@@
 We refer to such hybrid minimum
-lifetables as best practices lifetables.
+lifetables as best practices lifetables. \textit{eikemo2014} use the term ``best
+practices'' in a similar sense of an ideal mix of condtions, but do so via
+combining particular risk factors with their estimated mortality impacts to
+produce best case scenarios. The end effect is quite similar, to the extent that
+either estimation strategy shows what society could acheive given perfect
+diffusion of some defintion of best conditions. In the case of a best practices
+lifetable, we condition on mortality outcomes only, making the endeavor much
+simpler in practice.
 
@@
 converted to fractions. Fractions were then multiplied into the single-age
-all-cause death rates.
+all-cause death rates. The eleven causes of death considered in this abstract
+version are displayed in Table~\ref{tab:desc}, with their relative percentages.
+
+% latex table generated in R 3.1.2 by xtable 1.7-4 package
+% Wed Sep 23 17:46:55 2015
+\begin{table}[ht]
+\centering
+\caption{Eleven causes of death considered in the current study version,
+percentages.}
+\label{tab:desc}
+\begin{tabular}{rrr}
+ & Females & Males \\ 
+  \hline
+All other & 29.91 & 33.95 \\ 
+  Breast & 3.73 & 0.03 \\ 
+  Cardiovascular & 38.49 & 38.10 \\ 
+  Cerebrovascular & 10.50 & 6.84 \\ 
+  Lung & 3.55 & 6.50 \\ 
+  Other Cerebrovascular & 0.33 & 0.28 \\ 
+  Other Malignant Neoplasms & 11.01 & 9.83 \\ 
+  Other smoking & 0.57 & 1.36 \\ 
+  Prostate &  & 2.25 \\ 
+  Stomach & 0.67 & 0.87 \\ 
+  Uterus & 1.25 &  \\ 
+   \hline
+\end{tabular}
+\end{table}
 
@@
 male national average life expectancy. The respective rates of improvement for
-females have been 0.1470 and 0.1456.
+females have been 0.1470 and 0.1456. It is convenient and telling that the
+respective medium term average rates of improvement between national and
+best practices life expectancy match so closely, despite the best practices
+trend having a much more stable trajectory. This artifact may provide some
+support to conceiving of the best practices trend as indicative of where the
+U.S. is going in the coming decades.
 
@@
 
-\subsection{State differences by cause}
+If we take the mean rate of improvement for national or best practice trends,
+and compare it with the means of each series, we can roughly conclude that the
+best practices trend in year $t$ is a decent estimate of the national life
+expectancy in 30 or so years. This time lag is valid for the present series,
+but will vary depending on the number of cause groups that enter into the
+selection of minimum mortality rates. Clearly, if the population units were
+counties rather than states, the level of best practices life expectancy would
+increase even further, but we opine that cause-specific smoothing at the county
+level is more of a modelling gesture than sound practice, due to very low counts of deaths in
+most counties in the U.S. In the final paper we will aim to test the cause set
+proposed by \citet{vallin2008minimum} and compare it with other reasonable
+configurations for the U.S., and so recommend a rule of thumb grouping.
+
+\FloatBarrier
+\subsection*{State differences by cause}
 
 We decompose the state departures in life expectancy from the best practice
-trend in each year and sex, over both age and cause. In Figure~\ref{fig:dep}
+trend in each year and sex, over both age and cause. Part of the attraction of decomposing with respect to the best practices life expectancy is that contributions are always
+of the same sign, which makes interpretation of the heatmap more intuitive.
+Deeper shades of red indicate larger departures. In Figure~\ref{fig:dep}
 we aggregate over age and average within decades to summarize the total
@@
 based on the average for the years 2000-2004 only.} Major causes of death are in
-rows, and decades in columns.
-Part of the attraction of decomposing with respect to the best practices life expectancy is that contributions are always
-of the same sign, which makes interpretation of the heatmap more intuitive.
-Hotter colors indicate larger departures. For the case of cardiovascular disease
+rows, and decades are organized in columns.
+
+For the case of cardiovascular disease
 and other causes, grouping departures into the 2$+$ group blends out much
 interstate variation, since some contributions range as high as 4
-years.\footnote{This visual display may be significantly adapted when we add
-more causes to the decomposition. We will also do a similar small multiples
+years, but broad patterns are still visible.\footnote{This visual display may be
+significantly adapted when we add more causes to the decomposition. We will also do a similar small multiples
 display for grouped age patterns.}
@@
 state differences, trends are dominated by Cardivascular diseases and the large
-aggregate group of all other causes. The 11 original causes were grouped so as
-to make state patterns more visible.
+aggregate group of all other causes. The 11 original causes were grouped into
+five so as to make state patterns more visible. More commentary to come at a
+later date.
 
@@
 
+\section*{}
+
 \FloatBarrier

### 2015-09-23 BPLE/R/Descriptives.R

@@ -1,0 +1,22 @@
+
+# Author: tim
+###############################################################################
+
+if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
+	# if I'm on the laptop
+	setwd("/home/tim/git/BPLE/BPLE")
+} else {
+	# in that case I'm on Berkeley system, and other people in the dept can run this too
+	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
+}
+getwd()
+library(data.table)
+source("/home/tim/git/BPLE/BPLE/R/LTuniform.R")
+
+LTC <- local(get(load("Data/LTC1.Rdata")))
+head(LTC)
+hm <- tapply(LTC$mxcs2*LTC$Exposure, list(LTC$Cause, LTC$Sex), sum, na.rm=TRUE)
+hm <- as.data.frame(hm)
+hm$f[9] <- NA
+library(xtable)
+print(xtable(100*t(t(hm) / colSums(hm,na.rm=TRUE))))

### 2015-09-23 BPLE/R/e0trends.R

@@ -94,3 +94,8 @@ diff(range(bpf))/45;diff(c(use0f[1],use0f[46]))/46
 diff(c(use0m[1],use0m[46]))/46
 diff(c(use0f[1],use0f[46]))/46
 
+# average difference
+(mean(bpm) - mean(use0m)) / (diff(c(use0m[1],use0m[46]))/46)
+(mean(bpf) - mean(use0f)) / (diff(c(use0f[1],use0f[46]))/46)
+
+

### 2015-09-24 BPLE/Manuscript.tex
changes as of early afternoon, Thursday

@@
 \affil[1]{Max Planck Institute for Demographic Research}
-\affil[2]{universit{\'e} de Gen{\`e}ve}
-\affil[3]{Institut National d'{\'e}tudes D{\'e}mgraphiques}
+\affil[2]{Universit{\'e} de Gen{\`e}ve}
+\affil[3]{Institut national d'{\'e}tudes d{\'e}mographiques}
 \affil[4]{Department of Demography, University of California, Berkeley}
@@
 We calculate a preliminary series of best-possible lifetables for the United
-States from 1959-2004, defined as the age-specific aggregate of the lowest
+States from 1959 to 2004, defined as the age-specific aggregate of the lowest
 observed age-cause specific death rates among the 50
@@
 respectively. We argue that the US states best practices lifetable is a useful
-guage of mortality.
+gauge of future mortality trends.
 \ackn{The work reported in this manuscript was supported in part by the U.S.
@@
 future. Most past attempts to place limits on such projections were shown to be
-overly conservative within a short period of years \citep{oeppen2002broken}.
+overly conservative within a short period \citep{oeppen2002broken}.
 However, many simple mortality projections will send life expectancy scenarios
-to very high levels that for a given population may seem unimaginable. In some
+to very high levels that for a given population may seem unrealistic. In some
 cases, one seeks an external population that has already achieved a life
@@
 2012 life expectancy for Japanese females was 86, while for US females it was 81
-\citep{HMD}. If a US projection turns out to reach 86 we at least know that this
-is indeed possible.
+\citep{HMD}. If a US projection turns out to reach 86 we at least know that
+there exists some set of conditions in the world today that imply that this is
+indeed possible.
 
@@
 populations may one day achieve due to diffusion in practices, technology, and
-wellbeing \citep{vallin2010esperance}. The vanguard life expectancy is
-calculated based on mortality rates undifferentiated by cause of death, which
-may not provide an omnibus signal of what may eventually develop. While the force of
-mortality governs the lifetable, there is a substantive rationale to conceive of
+wellbeing \citep{vallin2010esperance}. The highest observed life expectancy in
+a given moment, which we term the ``vanguard'' life expectancy, is calculated
+based on mortality rates undifferentiated by cause of death.
+All cause mortality may not provide an omnibus signal of what may eventually develop.
+While the force of mortality governs the lifetable, there is a substantive rationale to conceive of
 this force as a composite of cause-specific forces of mortality.
@@
 We refer to such hybrid minimum
-lifetables as best practices lifetables. \textit{eikemo2014} use the term ``best
-practices'' in a similar sense of an ideal mix of condtions, but do so via
+lifetables as ``best practices'' lifetables. \textit{eikemo2014} use the term
+``best practices'' in a similar sense of an ideal mix of condtions, but do so via
 combining particular risk factors with their estimated mortality impacts to
@@
 calculated on the basis of its own territorial subdivisions, the 50 states plus
-the District of Columbia. This choice limits the sources of unaccounted for
-heterogeneity in international comparisons, and it also conveniently homogenizes
-data sources. To be clear, we do not necessarily propose using a trend in best
-practices mortality as a component of coherent mortality forecasts, and we do
+the District of Columbia. This choice limits the sources of unaccounted
+heterogeneity in international comparisons, and it also guarantees uniformity of
+data sources.
+To be clear, we do not necessarily propose using a trend in best practices mortality as a component of coherent mortality forecasts, and we do
 not undertake such an exercise. Instead, we view the U.S. best practices
@@
 All death count data come from the public NCHS mortality microdata files
-\citep{NCHSdata}. Population denominators come from a beta version of the U.S.
+\citep{NCHSdata}.\footnote{The current analysis is based on the public files
+with geographic information only up to 2004. We have bene granted access to
+geographic variables in a NCHS Research Data Center, and will extend analysis
+through 2013 for the final paper version.} Population denominators come from a
+beta version of the U.S.
 States project of the Human Mortality Database \citep{HMD}. Further processing of
@@
 the \texttt{MortalitySmooth} package \citep{GC2012} in \texttt{R}. The age-period
-matrix of death rates for each state and sex is smoothed over both margins,
-fractions are then recalculated, a multiplied into the smoothed all-cause death
-rates. Finally, we calculate lifetables for both states and best practices
-lifetables using HMD methodology.
+matrix of death rates for each state and sex is smoothed over both margins
+using a penalized b-spline. Cause fractions are then recalculated, and
+multiplied into the smoothed all-cause death rates. Finally, we calculate
+lifetables for each sex and state, as well as the best practices lifetable of
+each year using HMD methodology.
 
@@
 
+\subsection{Vanguard and best practices life expectancy}
+For the sake of clarity, we define vanguard life expectancy as the highest
+observed state life expectancy for each sex in a given year. Elsewhere, this
+quantity is called the record life expectancy \citep{oeppen2002broken}, or even
+the best practices life expectancy \citep{sanderson2004putting}, but we apply
+the term best practices in a different sense. To define the best practices
+mortality schedule, select the lowest (smoothed) death rate by age, $x$,
+and cause, $c$, from among all the states, $s$, $m(s,x,c)$, and sum over
+all causes, $C$, within ages to define the minimum mortality schedule,
+$m(x)^{min}$
+\begin{equation}
+m(x)^{min} = \sum _{c=1}^C min(m(s,x,c))
+\end{equation}
+
+The minimum mortality schedule is then used to calculate the best-practices life
+expectancy, and its cause-specific mortality pattern is used as the
+benchmark against which to decompose the mortality pattern of each state. While
+the best practices life expectancy is drawn from the diverse mortality
+experience of the entire USA, the vanguard life expectancy refers only to a
+particular state in a given year. We can say, for instance, that Minnesota was
+the vanguard for females in the year 1959, with a lif expectancy of 75.3, but
+that the best practices life expectancy in the same year was another 3 years
+higher at 78.3.
+
 \FloatBarrier
@@
 
-All results at this time are preliminary, but we can report broad trends at this
-time.\footnote{The final results will differ in the following ways. At this time we have
-calculated results using 11 causes of death that were previously used in a different research project, and this will
-change to 30 causes to be consistent and comparable with \citet{vallin2008minimum}. Including
+All results at this time are preliminary, but we can report broad trends.\footnote{The final results will differ in the following ways. At this time we have
+calculated results using 11 causes of death that were previously from a
+different research project, and this will change to 30 causes to be consistent and comparable with \citet{vallin2008minimum}. Including
 more causes will increase the best practices life expectancy level. We will also
@@
 produce results through 2004, because geographic information is suppressed in
-the microdata for later years. This information is available in secure Research
-Data Centers, where we will produce cause fractions and all-cause lifetables for
+the public microdata for later years. This information is available in secure
+Research Data Centers, where we will produce cause fractions and all-cause lifetables for
 the most recent decade.}
@@
 over the 45 years considered, except for males in the District of Columbia,
-which has been recovering strongly since the 1990s, due both to secular
+which have been recovering strongly since the 1990s, due both to secular
 improvements and compositional change. The US aggregate trend is bouncier
@@
 state life expectancy for males is Hawaii until the two most recent years of
-observation (2003-2004), where Minnesota has taken the lead. For females, the
-Minnesota is vanguard until 1963, followed by North Dakota for five years, and
-then Hawaii for the rest of the series until 2004. The vanguard trend for
-females has the same overall shape as the cluster of individual state trends,
-but for males the time trend in vanguard, ergo Hawaiian, life expectancy was
+observation (2003-2004), where Minnesota has taken the lead. For females,
+Minnesota was vanguard until 1963, followed by North Dakota for five years, and
+then Hawaii for the rest of the period until 2004. The vanguard trend for
+females has the same overall shape as the cluster of individual state trends, but for males the time trend in vanguard, ergo Hawaiian, life expectancy was
 different from the bulk of states.
@@
 practices trend is on average 2 years higher than the vanguard trend, and its
-progress is steadier than either the vanguard or national trends, as well as
-most individual states. The best practices trend has increased each year for
+progress is steadier than either the vanguard, national, or average state
+trends, as well as most individual states. The best practices trend has increased each year for
 males and females, though faster for males. The average annual improvement for
@@
 \begin{figure}[t!]
-\caption{Trends in US, vanguard, best practices, and state life expectancies
-($e(0)$), 1959-2004.}
+\caption{Trends in US, vanguard, best practices, and state life expectancies at
+birth ($e(0)$), 1959-2004.}
 \label{fig:e0}
@@
 trend in each year and sex, over both age and cause. Part of the attraction of decomposing with respect to the best practices life expectancy is that contributions are always
-of the same sign, which makes interpretation of the heatmap more intuitive.
-Deeper shades of red indicate larger departures. In Figure~\ref{fig:dep}
-we aggregate over age and average within decades to summarize the total
-contribution to the departure from best practices.\footnote{The 2000s trend is
-based on the average for the years 2000-2004 only.} Major causes of death are in
+of the same sign, which makes interpretation of the heatmaps more intuitive.
+Deeper shades of red indicate larger departures, i.e., larger $e(0)$
+shortcomings.
+In Figure~\ref{fig:dep} we represent the average $e(0)$ shortcoming from each
+cause and state, and average within decades to summarize the total contribution
+to the departure from best practices.\footnote{The 2000s trend is based on the average for the years 2000-2004 only.} Major causes of death are in
 rows, and decades are organized in columns.
+
+In these heat maps, states are represented as identically-sized square cells,
+while preserving rough spatial relationships from a standard map
+projection.\footnote{This configuration is inspired by a tile grid map on the
+NPR blog \citep{NPRsquares}, which
+itself was inspired by various commonly used maps in the New York Times,
+Washington Post, and others. DeBelius notes that grid panels are not meant for
+displaying population quantities, since each state has a different population
+size. Since lifetable quantities are purged of population size and
+structure, the visual instrument is adequate, and the equalization of state
+size and form adds legibility to our results.} For orientation, we present a
+state guide to the heatmaps in Figure~\ref{fig:heatguide}.
+\begin{figure}
+\centering
+\caption{A guide to tiled heatmaps of U.S. states.}
+\label{fig:heatguide}
+\includegraphics[scale=.3]{Figures/StatesDiagram.pdf}
+\end{figure}
 
@@
 
-\begin{figure}[t!]
-\caption{State departures from vanguard $e(0)$ by large cause groups,
-1959-2004.}
-\label{fig:dep}
+\begin{figure}
 \centering
-\begin{subfigure}[b]{\textwidth}
+\caption{U.S. males, state departures from best practices $e(0)$ by large cause
+groups, 1959-2004.}
+\label{fig:depm}
+\includegraphics[scale=0.7]{Figures/StatesDecadesM.pdf}
+\end{figure}
+
+\begin{figure}
 \centering
-\caption{Males}
-\label{fig:depm}
-\includegraphics[scale=0.5]{Figures/StatesDecadesM.pdf}
-\end{subfigure}
-\\
-\begin{subfigure}[b]{\textwidth}
-\centering
-\caption{Females}
+\caption{U.S. feales, state departures from best practices $e(0)$ by large cause
+groups, 1959-2004.}
 \label{fig:depf}
-\includegraphics[scale=0.5]{Figures/StatesDecadesF.pdf}
-\end{subfigure}
+\includegraphics[scale=0.7]{Figures/StatesDecadesF.pdf}
 \end{figure}

### 2015-09-24 BPLE/R/StatesDiagram.R

@@ -1,0 +1,38 @@
+# TODO: Add comment
+# 
+# Author: tim
+###############################################################################
+if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
+	# if I'm on the laptop
+	setwd("/home/tim/git/BPLE/BPLE")
+} else {
+	# in that case I'm on Berkeley system, and other people in the dept can run this too
+	setwd(paste0("/data/commons/",system("whoami",intern=TRUE),"/git/BPLE/BPLE"))
+}
+
+ST <- matrix(c("AK",rep("",10),"ME",
+		rep("",10),"VT","NH",
+		"","WA","ID","MT","ND","MN","IL","WI","MI","NY","RI","MS",
+		"","OR","NV","WY","SD","IA","IN","OH","PA","NJ","CT","",
+		"","CA","UT","CO","NB","MO","KY","WV","VA","MD","DL","",
+		"","","AZ","NM","KS","AR","TN","NC","SC","DC","","",
+		"","","","","OK","LA","MS","AL","GA","","","",
+		"HA","","","","TX","","","","","FL","",""),ncol = 12, nrow = 8,byrow=TRUE)
+
+ST[ST == ""] <- NA
+Color <- matrix(NA,nrow=nrow(ST),ncol=ncol(ST))
+Color[!is.na(ST[8:1,])] <- gray(.5)
+
+x1 <- col(ST) - 1
+x2 <- col(ST)
+y1 <- row(ST) - 1
+y2 <- row(ST)
+
+#coords <- centercoords(x1,x2,y1,y2)
+
+pdf("Figures/StatesDiagram.pdf",width=12,height=9)
+par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
+plot(NULL, xlim = c(0,12),ylim=c(0,9),xlab = "", ylab = "", type= "n", axes = FALSE,asp=1)
+rect(x1,y1,x2,y2,border="white",col = Color)
+text((x1+x2)/2,(y1+y2)/2,ST[8:1,],col="white",cex=2)
+dev.off()

### 2015-09-24 BPLE/R/e0trends.R

@@ -7,6 +7,7 @@ if (system("hostname",intern=TRUE) %in% c("triffe-N80Vm", "tim-ThinkPad-L440")){
 }
 getwd()
 library(data.table)
+library(reshape2)
 source("/home/tim/git/BPLE/BPLE/R/LTuniform.R")
 
 LTC <- local(get(load("Data/LTC1.Rdata")))
@@ -97,5 +98,13 @@ diff(c(use0f[1],use0f[46]))/46
 # average difference
 (mean(bpm) - mean(use0m)) / (diff(c(use0m[1],use0m[46]))/46)
 (mean(bpf) - mean(use0f)) / (diff(c(use0f[1],use0f[46]))/46)
+
+bpf[1]-e0fs["MN","1959"]
+
+summary(lm(colMeans(e0ms)~yrs))$r.squared
+summary(lm(colMeans(e0fs)~yrs))$r.squared
+
+colMeans(e0fs)
+
 
 

### 2015-09-24 BPLE/Manuscript.tex
added in most of Magali's changes, now to look at Adrien's email

@@
 
+There are limitations to the interpretation of a best practices trend, as
+defined here. The mix of causes within each age, and the resulting age pattern
+for each cause may not necessarily reflect a possible or likely mortality
+pattern. For instance, diseases may compensate each other, especially for
+diagnostic and reporting reasons. For instance, for decedents that suffered
+from both condition A and condition B, one state might be more likely to
+declare A as the underlying cause of death, while another state would more
+likely declare B to be as the underlying cause. A situation like this
+would tend to inflate the best practices $e(0)$. The best practices $e(0)$ may
+be underestimate in instances where reductions in cause A are expected to lead
+to reductions in cause B (e.g., pneumonia and cardiovascular disease.)
 \FloatBarrier
@@
 
+For the sake of comparison with the presentlt calculated best practices series.
+The $e(0)$ value in 2004 for males and females was 80 and 84.9, respectively. In
+the same year, the observed all-cause $e(0)$ in Japan were 78.6 for males and
+85.5 for females. That is, the U.S. best practices trend is ahead of observed
+life expectancy for Japanese males, and behind for Japanese females. We may also
+compare with the official $e(0)$ projections of the U.S. Social Security
+Administration (SSA). The SSA projects projects male $e(0)$ passing 80 in the
+year 2058, and female $e(0)$
+passing 84.9 in the year 2075 \citep{SSA2005}. These projections are a full 54
+and 71 years after 2004, in both cases much farther in the future than a simple extrapolation
+of the best practices trend would predict, circa 2034.
+
 \FloatBarrier
@@
 
-We decompose the state departures in life expectancy from the best practice
-trend in each year and sex, over both age and cause. Part of the attraction of decomposing with respect to the best practices life expectancy is that contributions are always
-of the same sign, which makes interpretation of the heatmaps more intuitive.
-Deeper shades of red indicate larger departures, i.e., larger $e(0)$
-shortcomings.
-In Figure~\ref{fig:dep} we represent the average $e(0)$ shortcoming from each
-cause and state, and average within decades to summarize the total contribution
-to the departure from best practices.\footnote{The 2000s trend is based on the average for the years 2000-2004 only.} Major causes of death are in
-rows, and decades are organized in columns.
-
-In these heat maps, states are represented as identically-sized square cells,
-while preserving rough spatial relationships from a standard map
-projection.\footnote{This configuration is inspired by a tile grid map on the
+Since the trend in the best practices $e(0)$ is so stable over time, it
+is also a convenient moving benchmark against which to compare state patterns
+--- a kind of potential mortality for each state. We
+decompose the state departures in life expectancy from the best practice trend in each year and sex, over both age and cause.
+Part of the attraction of decomposing with respect to the best practices life expectancy is that contributions are always of the same sign, which makes heatmaps more intuitive than would otherwise be the case.\footnote{Typically age and/or cause-specific contributions to a difference in $e(0)$ are represented in stacked bar charts, which in the present case would not be a practical way to distill information.} We therefore summarize state trends in a series of heatmaps representing the 50 states plus the District of Columbia.
+In these heat maps, states are represented as identically-sized square cells, while preserving rough spatial relationships from a standard map projection.
+For orientation, we present a state guide to the heatmaps in Figure~\ref{fig:heatguide}.\footnote{This configuration is inspired by a tile grid map on the
 NPR blog \citep{NPRsquares}, which
@@
 structure, the visual instrument is adequate, and the equalization of state
-size and form adds legibility to our results.} For orientation, we present a
-state guide to the heatmaps in Figure~\ref{fig:heatguide}.
+size and form adds legibility to our results.} Deeper shades of red
+indicate larger departures, i.e., larger $e(0)$ shortcomings.
+
 \begin{figure}
@@
 
-For the case of cardiovascular disease
-and other causes, grouping departures into the 2$+$ group blends out much
+In Figures~\ref{fig:depm} and \ref{fig:depf} we represent the average $e(0)$
+shortcoming from each cause and state, and average within decades to summarize the total contribution
+to the departure from best practices.\footnote{The 2000s results are based on
+the average for the years 2000-2004 only.} Major causes of death are in rows,
+and decades are organized in columns. For the case of cardiovascular disease and other causes, grouping departures into the 2$+$ group blends out much
 interstate variation, since some contributions range as high as 4
@@
 
+Glancing over the lung diseases and cancer rows of the maps, it is evident that
+the District of Columbia had high excess mortality for these causes over all
+five decades, peaking in the 1980s. Lung diseases display a clear mortality
+basin in the Southern states for males, although the absolute maginitude of these departures from ideal mortality has diminished over time. The same mortality basin occurs in
+cariovascular causes over all five decades, but is clearly visible in this
+display only since the 1990's, due to the 2$+$ grouping.
+
 Since the cause groupings available at the time of submission are not the final groupings, and
@@
 aggregate group of all other causes. The 11 original causes were grouped into
-five so as to make state patterns more visible. More commentary to come at a
-later date.
+five so as to make state patterns more visible. These results will be commented
+at greater length when the final results are prepared.
 
@@
 
-\section*{}
-
 \FloatBarrier
+\section*{Discussion}
+In this paper, we have given an intuitive rationale for calculating a best
+practices trend in the United States. First, the best practices trend in life
+expectancy at birth can be interpreted as the mortality potential present in the
+United States at a given point in time. It is an indication of the mortality
+conditions that would come to be if local conditions, practices,
+technologies, and behaviors were perfectly diffused from those places with the
+best age and cause-specific performance in the year. While it is difficult to
+imagine that pressing a magic button to activate this perfect cocktail of
+mortality risk factors would actually bring about this minimum mortality
+instantaneously--- for instance, cumulative risk factors in population
+stocks that are distributed unevenly over space would stymie such a thought
+experiment--- we do observe catch-ups with past best practices values for many states, and for the
+U.S. national $e(0)$, on average after a period of 30 years. It remains to be
+seen how this lag scales to the number of causes ocnsidered and size and
+distribution of subpopulations. At this time, we are rather attracted to the
+idea of a period weathervane that points us to where period mortality is going.
 

### 2015-09-24 BPLE/R/StatesSquares.R

@@ -172,3 +172,18 @@ rect(xat,yat[-length(yat)],xat + .4,yat[-1],col=ramp(length(lbrks)-1))
 text(xat+.4,yat,paste0(lbrks,c("","","","","+")),pos=4)
 text(xat+.2,max(yat)+.3,"Years e(0)")
 dev.off()
+
+
+colMeans(F1960s)
+colMeans(F1970s)
+colMeans(F1980s)
+colMeans(F1990s)
+colMeans(F2000s)
+
+hist(F1960s[,"Cancer"])
+hist(F2000s[,"Cancer"])
+
+
+
+
+

### 2015-09-24 BPLE/R/e0trends.R

@@ -106,5 +106,5 @@ summary(lm(colMeans(e0fs)~yrs))$r.squared
 
 colMeans(e0fs)
 
-
-
+bpm
+bpf

### 2015-09-24 BPLE/Manuscript.tex
.

@@
 
-\subsection{Vanguard and best practices life expectancy}
+\subsection*{Vanguard and best practices life expectancy}
 For the sake of clarity, we define vanguard life expectancy as the highest

### 2015-09-24 BPLE/build.R

@@ -5,4 +5,7 @@ system("bibtex Manuscript.aux") # to generate .bbl
 system("pdflatex Manuscript.tex Manuscript.pdf")
 system("pdflatex Manuscript.tex Manuscript.pdf")
 
-file.copy("Manuscript.pdf","/home/tim/Dropbox/BPLE/Manuscript/Manuscript.pdf",overwrite=TRUE)
+file.copy("Manuscript.pdf","/home/tim/Dropbox/BPLE/Manuscript/Manuscript.pdf",
+		overwrite=TRUE)
+
+

### 2015-09-24 BPLE/Manuscript.tex
.

@@
 
-\title{A best practices composite lifetable for US states}
+\title{A best practices composite lifetable for the United States}
 
@@
 We calculate a preliminary series of best-possible lifetables for the United
-States from 1959 to 2004, defined as the age-specific aggregate of the lowest
-observed age-cause specific death rates among the 50
-states and the District of Columbia for each year. This synthetic best practices lifetable shows a gradual
-increasing trend over the period, on average 1.9 and 2.2 years higher
-than the highest state life expectancy in each year for males and females,
-respectively. We argue that the US states best practices lifetable is a useful
-gauge of future mortality trends.
+States from 1959 to 2004, defined on the basis of the age-specific aggregate of
+the lowest observed age-cause specific death rates among the 50
+states and the District of Columbia for each year. The resulting highest-possible life expectancy shows a gradual increasing trend over the period, on
+average 1.9 and 2.2 years higher than the highest state life expectancy in each year for males and females, respectively. We argue that the US states best practices lifetable is a useful
+gauge of future mortality trends and a good benchmark against which to compare
+state mortality patterns.
 \ackn{The work reported in this manuscript was supported in part by the U.S.
@@
 We refer to such hybrid minimum
-lifetables as ``best practices'' lifetables. \textit{eikemo2014} use the term
+lifetables as ``best practices'' lifetables. \citet{eikemo2014} use the term
 ``best practices'' in a similar sense of an ideal mix of condtions, but do so via

### 2015-09-25 BPLE/Manuscript.tex
fixed p spline wording
accounted for Celeste's edits, ran spell-check!

@@
 lifetables as ``best practices'' lifetables. \citet{eikemo2014} use the term
-``best practices'' in a similar sense of an ideal mix of condtions, but do so via
-combining particular risk factors with their estimated mortality impacts to
+``best practices'' in a similar sense of an ideal mix of conditions, but do so
+via combining particular risk factors with their estimated mortality impacts to
 produce best case scenarios. The end effect is quite similar, to the extent that
-either estimation strategy shows what society could acheive given perfect
-diffusion of some defintion of best conditions. In the case of a best practices
+either estimation strategy shows what society could achieve given perfect
+diffusion of some definition of best conditions. In the case of a best practices
 lifetable, we condition on mortality outcomes only, making the endeavor much
@@
 National populations are heterogeneous with respect to many factors, both
-observed and unobserved, that affect cause-spectific mortality rates. This does
+observed and unobserved, that affect cause-specific mortality rates. This does
 not make international mortality comparisons futile, but it may make perfect
@@
 matrix of death rates for each state and sex is smoothed over both margins
-using a penalized b-spline. Cause fractions are then recalculated, and
+using a p-spline. Cause fractions are then recalculated, and
 multiplied into the smoothed all-cause death rates. Finally, we calculate
@@
 the term best practices in a different sense. To define the best practices
-mortality schedule, select the lowest (smoothed) death rate by age, $x$,
-and cause, $c$, from among all the states, $s$, $m(s,x,c)$, and sum over
+mortality schedule, for each year and sex separately select the lowest
+(smoothed) death rate by age, $x$, and cause, $c$, from among all the states, $s$, $m(s,x,c)$, and sum over
 all causes, $C$, within ages to define the minimum mortality schedule,
@@
 which have been recovering strongly since the 1990s, due both to secular
-improvements and compositional change. The US aggregate trend is bouncier
-because death counts were not smoothed, and its rank position among states
-varies due in part also to compositional changes between states. The vanguard
+improvements and compositional change. The US aggregate trend is less smooth
+because national death counts were not smoothed, and its rank position among
+states varies due in part also to compositional changes between states. The vanguard
 state life expectancy for males is Hawaii until the two most recent years of
@@
 proposed by \citet{vallin2008minimum} and compare it with other reasonable
-configurations for the U.S., and so recommend a rule of thumb grouping.
+configurations for the U.S., and so recommend a suitable cause grouping.
 
-For the sake of comparison with the presentlt calculated best practices series.
+For the sake of comparison with the presently calculated best practices series.
 The $e(0)$ value in 2004 for males and females was 80 and 84.9, respectively. In
@@
 five decades, peaking in the 1980s. Lung diseases display a clear mortality
-basin in the Southern states for males, although the absolute maginitude of these departures from ideal mortality has diminished over time. The same mortality basin occurs in
-cariovascular causes over all five decades, but is clearly visible in this
+basin in the Southern states for males, although the absolute magnitude of these departures from ideal mortality has diminished over time. The same mortality basin occurs in
+cardiovascular causes over all five decades, but is clearly visible in this
 display only since the 1990's, due to the 2$+$ grouping.
@@
 because the causes included in the study contribute such varying magnitudes to
-state differences, trends are dominated by Cardivascular diseases and the large
+state differences, trends are dominated by cardiovascular diseases and the large
 aggregate group of all other causes. The 11 original causes were grouped into
@@
 U.S. national $e(0)$, on average after a period of 30 years. It remains to be
-seen how this lag scales to the number of causes ocnsidered and size and
+seen how this lag scales to the number of causes considered and size and
 distribution of subpopulations. At this time, we are rather attracted to the

### 2016-02-28 BPLE/Manuscript.tex
this project needs to be kickstarted again for upcoming lab talk. Soon to become focus

@@
 %\newcommand\Cell[1]{{\raisebox{-0.05in}{\includegraphics[height=.2in,width=.2in]{Figures/ColorCodes/\expandafter#1}}}}  
-
+\defcitealias{HMD}{HMD}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\defcitealias{HMD}{HMD}
+%\defcitealias{HMD}{HMD}
 
@@
 the lowest observed age-cause specific death rates among the 50
-states and the District of Columbia for each year. The resulting highest-possible life expectancy shows a gradual increasing trend over the period, on
+states and the District of Columbia. The resulting highest-possible life expectancy shows a gradual increasing trend over the period, on
 average 1.9 and 2.2 years higher than the highest state life expectancy in each year for males and females, respectively. We argue that the US states best practices lifetable is a useful
@@
 expectancy may not reflect the best mortality possible given current conditions
-over a set of populations. For instance, the lowest rate of suicide and heart
+in a set of populations. For instance, the lowest rate of suicide and heart
 disease at age 45 will not necessarily be found in the same population. As an
@@
 beta version of the U.S.
-States project of the Human Mortality Database \citep{HMD}. Further processing of
+States project of the \citet{HMD}. Further processing of
 population counts into exposures is done according to the HMD Methods
@@
 One caveat of calculating a best practices lifetable is that stochastic zeros
-occur in particular ages and causes, especially among small populations. The
+occur in particular ages and causes, especially in small populations. The
 exercise of selecting the minimum death rate for a given age and cause will tend
@@
 life expectancy. We therefore smooth all death rates using
-the \texttt{MortalitySmooth} package \citep{GC2012} in \texttt{R}. The age-period
+the method proposed by \citep{GC2012}. The age-period
 matrix of death rates for each state and sex is smoothed over both margins
@@
 Results are summarized by ages, causes and states and can be interpreted
-directly as the the total lag in years of life expectancy due to excess
-mortality (anything greater than the minimum) in the given margin.\footnote{In
-the final version we may opt to further separate trends from initial
-differences using the decomposition method recently proposed by
-\citet{DimaDecomp2014}.}
+directly as the the total gap in years of life expectancy due to excess
+mortality (anything greater than the minimum) in the given margin.
+%\footnote{In
+%the final version we may opt to further separate trends from initial
+%differences using the decomposition method recently proposed by
+%\citet{DimaDecomp2014}.}
 
 \subsection*{Vanguard and best practices life expectancy}
-For the sake of clarity, we define vanguard life expectancy as the highest
+We define vanguard life expectancy as the highest
 observed state life expectancy for each sex in a given year. Elsewhere, this
@@
 particular state in a given year. We can say, for instance, that Minnesota was
-the vanguard for females in the year 1959, with a lif expectancy of 75.3, but
-that the best practices life expectancy in the same year was another 3 years
+the vanguard for females in the year 1959, with a life expectancy of 75.3, but
+that the best practices life expectancy in the same year was 3 years
 higher at 78.3.
@@
 for each cause may not necessarily reflect a possible or likely mortality
-pattern. For instance, diseases may compensate each other, especially for
+pattern. Diseases may compensate each other, especially for
 diagnostic and reporting reasons. For instance, for decedents that suffered
 from both condition A and condition B, one state might be more likely to
-declare A as the underlying cause of death, while another state would more
-likely declare B to be as the underlying cause. A situation like this
-would tend to inflate the best practices $e(0)$. The best practices $e(0)$ may
-be underestimate in instances where reductions in cause A are expected to lead
+declare A as the underlying cause of death, while another state would on
+average more likely declare B to be as the underlying cause. A situation like
+this would tend to inflate the best practices $e(0)$. The best practices $e(0)$ may
+be underestimated in instances where reductions in cause A are expected to lead
 to reductions in cause B (e.g., pneumonia and cardiovascular disease.)
@@
 male national average life expectancy. The respective rates of improvement for
-females have been 0.1470 and 0.1456. It is convenient and telling that the
+females have been 0.1470 and 0.1456. It is convenient and revealing that the
 respective medium term average rates of improvement between national and
@@
 but will vary depending on the number of cause groups that enter into the
-selection of minimum mortality rates. Clearly, if the population units were
+selection of minimum mortality rates. If the population units were
 counties rather than states, the level of best practices life expectancy would
-increase even further, but we opine that cause-specific smoothing at the county
-level is more of a modelling gesture than sound practice, due to very low counts of deaths in
-most counties in the U.S. In the final paper we will aim to test the cause set
-proposed by \citet{vallin2008minimum} and compare it with other reasonable
-configurations for the U.S., and so recommend a suitable cause grouping.
+increase even further, but this exercise would be more challenging, due to very
+low counts of deaths in most counties in the U.S. Increasing the number of
+causes of death would similarly increase estimates would increase in a like
+manner, as would subdividing the population in other categories, such as race
+or income. In the final paper we will aim to test the cause set proposed by
+\citet{vallin2008minimum} and compare it with other reasonable configurations for the U.S., and so recommend a suitable cause grouping.
 
@@
 and 71 years after 2004, in both cases much farther in the future than a simple extrapolation
-of the best practices trend would predict, circa 2034.
+of the best practices trend would predict, around 2034. If this seems unlikely,
+recall that the rate components necessary to acheive this improvement are
+already observed in the United States!
+
+Note that both male and female best practice trends increased unabated through
+the 1960s, where life expectancy tended to stagnate at the national and state
+levels (cit). Had we forecast mortality in the following decades based on data
+only through the 1960s history tells us we would underestimate future life
+expectancy. Even in the 1960s the best practices trend would have told
+us to aim forecasts higher.
 

### 2016-02-28 BPLE/R/StatesHeat.R

@@ -2,6 +2,26 @@
 ###############################################################################
 source("R/colourtools.R")
 
+#NYT <- read.csv("/home/tim/git/BPLE/BPLE/Data/NYTgridUS.csv")
+#
+#recvec <- 0:7
+#names(recvec) <- 7:0
+#NYT$y <- recvec[as.character(NYT$y)]
+
+#class(NYT) <-"gridUS"
+plot.gridUS <- function(X,...){
+	plot(NULL, type = "n", 
+			axes=FALSE,
+			xlab="",
+			ylab="",
+			asp=1,
+			xlim=range(X$x)+c(0,1),
+			ylim=range(X$y)+c(0,1))
+	rect(X$x,X$y,X$x+1,X$y+1,col=gray(.9),...)
+	text(X$x+.5,X$y+.5,X$key)
+}
+save(NYT,file="/home/tim/git/BPLE/BPLE/Data/NYTgridUS.Rdata")
+
 # custom, for artisanal panel plots
 centercoords <- function(x1,x2,y1,y2){
 	xrange <- range(c(x1,x2))

### 2016-03-06 BPLE/Manuscript.tex
started working in new cod fractions

@@
 
-\title{A best practices composite lifetable for the United States}
+\title{A low-mortality benchmark for the United States}
 
@@
 states and the District of Columbia. The resulting highest-possible life expectancy shows a gradual increasing trend over the period, on
-average 1.9 and 2.2 years higher than the highest state life expectancy in each year for males and females, respectively. We argue that the US states best practices lifetable is a useful
-gauge of future mortality trends and a good benchmark against which to compare
-state mortality patterns.
+average 1.9 and 2.2 years higher than the highest state life expectancy in each
+year for males and females, respectively. We argue that the US states
+low-morality lifetable is a useful gauge of future mortality trends. We apply
+the low-mortality as a benchmark to compare state mortality
+patterns and quantify the potential for immediately acheivable improvement.
 \ackn{The work reported in this manuscript was supported in part by the U.S.
@@
 
-The level and timing of cause-specific responses to particular
-technological or well-being improvements varies between populations. Therefore,
-the lowest observed all-cause force of mortality is not necessarily composed of
-the lowest cause-specific forces of mortality, which means that the vanguard life
-expectancy may not reflect the best mortality possible given current conditions
-in a set of populations. For instance, the lowest rate of suicide and heart
+The lowest observed all-cause force of mortality is not
+necessarily composed of the lowest cause-specific forces of mortality. This
+means that the vanguard life expectancy may not reflect the best mortality possible given current conditions in a set of populations. For instance, the lowest rate of suicide and heart
 disease at age 45 will not necessarily be found in the same population. As an
@@
 National populations are heterogeneous with respect to many factors, both
-observed and unobserved, that affect cause-specific mortality rates. This does
+observed and unobserved, that affect cause-specific mortality rates. The level and timing of cause-specific responses to particular
+technological or well-being improvements varies between populations. This does
 not make international mortality comparisons futile, but it may make perfect
@@
 \section*{Data and methods}
-All death count data come from the public NCHS mortality microdata files
-\citep{NCHSdata}.\footnote{The current analysis is based on the public files
-with geographic information only up to 2004. We have bene granted access to
-geographic variables in a NCHS Research Data Center, and will extend analysis
-through 2013 for the final paper version.} Population denominators come from a
-beta version of the U.S.
-States project of the \citet{HMD}. Further processing of
-population counts into exposures is done according to the HMD Methods
-Protocol, and all-cause lifetables were calculated by single ages \citep{HMDMP}.
-Death counts for causes were aggregated into quinquennial age groups by year and sex for eleven large groups that minimize disruptions between ICD transitions, and then
-converted to fractions. Fractions were then multiplied into the single-age
-all-cause death rates. The eleven causes of death considered in this abstract
+All death count data come from the restricted NCHS mortality microdata files
+\citep{NCHSdata}.\footnote{We will extend analysis
+through 2013 for the final paper version.} Single-age population denominators
+and complete state lifetables come from a beta version of the U.S.
+States project of the \citet{HMD}. Death counts by causes of death are first
+grouped into 30 large cause groups and then converted to fractions in abridged
+age groups, $[0,1)$, $[1,4)$, $[5,9)$, \ldots, $100+$ \citep{HMDMP}. Fractions
+in abridged age groups are then multiplied into single-age all-cause death
+rates to produce a first estimate of single-age cause-specific death rates.
+The 30 causes of death considered in this abstract
 version are displayed in Table~\ref{tab:desc}, with their relative percentages.
@@
 % Wed Sep 23 17:46:55 2015
-\begin{table}[ht]
-\centering
-\caption{Eleven causes of death considered in the current study version,
-percentages.}
-\label{tab:desc}
-\begin{tabular}{rrr}
- & Females & Males \\ 
-  \hline
-All other & 29.91 & 33.95 \\ 
-  Breast & 3.73 & 0.03 \\ 
-  Cardiovascular & 38.49 & 38.10 \\ 
-  Cerebrovascular & 10.50 & 6.84 \\ 
-  Lung & 3.55 & 6.50 \\ 
-  Other Cerebrovascular & 0.33 & 0.28 \\ 
-  Other Malignant Neoplasms & 11.01 & 9.83 \\ 
-  Other smoking & 0.57 & 1.36 \\ 
-  Prostate &  & 2.25 \\ 
-  Stomach & 0.67 & 0.87 \\ 
-  Uterus & 1.25 &  \\ 
-   \hline
-\end{tabular}
-\end{table}
+%\begin{table}[ht]
+%\centering
+%\caption{Eleven causes of death considered in the current study version,
+%percentages.}
+%\label{tab:desc}
+%\begin{tabular}{rrr}
+% & Females & Males \\ 
+%  \hline
+%All other & 29.91 & 33.95 \\ 
+%  Breast & 3.73 & 0.03 \\ 
+%  Cardiovascular & 38.49 & 38.10 \\ 
+%  Cerebrovascular & 10.50 & 6.84 \\ 
+%  Lung & 3.55 & 6.50 \\ 
+%  Other Cerebrovascular & 0.33 & 0.28 \\ 
+%  Other Malignant Neoplasms & 11.01 & 9.83 \\ 
+%  Other smoking & 0.57 & 1.36 \\ 
+%  Prostate &  & 2.25 \\ 
+%  Stomach & 0.67 & 0.87 \\ 
+%  Uterus & 1.25 &  \\ 
+ %  \hline
+%\end{tabular}
+%\end{table}
 

### 2016-03-06 BPLE/R/DataPrep.R

@@ -13,29 +13,10 @@ library(reshape2)
 library(magrittr)
 library(data.table)
 library(MortalitySmooth)
+library(HMDHFDplus)
 
-LT 			<- local(get(load("Data/MilaStatesSingleAge.Rdata")))
-
-# clean mxc, reshape
-mxc 		<- local(get(load("Data/mxcod.Rdata")))
-mxc 		<- melt(mxc, varnames = c("State","Year","Sex","Age","Cause"), value.name = "Mxc")
-mxc$State 	<- as.character(mxc$State)
-mxc$Sex   	<- as.character(mxc$Sex)
-mxc$Age   	<- as.character(mxc$Age)
-mxc$Cause 	<- as.character(mxc$Cause)
-
-# fix age
-# I rather dislike this little operation,
-# but more legible with pipes
-mxc$Age %>% 
-		strsplit(split="-") %>% 
-		lapply("[[",1) %>% 
-		unlist %>% 
-		gsub(pattern="Over ", replacement = "") %>% 
-		as.integer -> mxc$Age
-mxc$Sex <- ifelse(mxc$Sex == "male","m","f")
-
-# convert mxc to proportions
+# functions used here:
+# convert mxc to proportions ( now imported as such, as fractions)
 makeprop <- function(Mxc){
 	prop <- Mxc 
 	if (sum(prop) > 0){
@@ -43,33 +24,24 @@ makeprop <- function(Mxc){
 	}
 	prop
 }
-mxc <- data.table(mxc)
-mxc[,prop := makeprop(Mxc), by = list(Year, State, Sex, Age)]
-
-###################################################
-# apply proportions to single age mx 
-LT <- LT[LT$Year <=2004, ]
-head(LT)
-sum(LT$Mx==0 & !is.na(LT$Mx))
-
-LT$Dxhat <- LT$Exposure * LT$mx
+# we smooth all-cause mortality 2d.
 # given a matrix od DX and EX, lets smooth rates!
 sm.mat <- function(DX, EX){
-	ages  		<- 0:110
-	years 		<- 1959:2004
+	ages  		<- as.integer(rownames(DX))
+	years 		<- as.integer(colnames(DX))
 	W     		<- EX
 	W[W > 0] 	<- 1
 	W[DX < 1 & W == 1] 	<- .1
 	fit   <- Mort2Dsmooth(
-			     x = ages, 
-				 y = years, 
-				 Z = DX,
-			     offset = log(EX), 
-				 W = W,
[…]

### 2016-03-14 BPLE/R/DataPrep.R
.

@@ -70,6 +70,7 @@ sm.chunk.cause <- function(.SD){
 # now processing.
 
 LT 			<- local(get(load("Data/MilaStatesSingleAge.Rdata")))
+
 # fractions sent from Magali Mar 2, 2016.
 # 1959-2013 (now go beyond Mila's lifetables by 3 years
 frac 		<- local(get(load("Data/cod.fractions.Rdata")))
@@ -90,6 +91,17 @@ frac$Sex <- ifelse(frac$Sex == "M","m","f")
 # now with causes through 2013,
 # we need to cut off at 2010
 frac <- frac[frac$Year <= 2010, ]
+
+# this step added for memory management.
+# each state to be processed separately.
+# I don't like saving intermediate files,
+# but this requires > 8Gb memory otherwise
+states <- unique(frac$State)
+for (st in states){
+	ST <- frac[frac$State == st, ]
+	save(ST, file = paste0("Data/StateChunks/frac/",st,".Rdata"))
+}
+rm(frac);gc()
 
 # lots of zeros
 sum(LT$Mx==0 & !is.na(LT$Mx))
@@ -164,3 +176,5 @@ LTC$mxcs2 <- LTC$prop2 * LTC$mxs
 # so depending on whether the result is smooth or not, may be due for a second round of smoothing
 # within cause. Or maybe we accept this? Probably enough juice for now.
 save(LTC, file = "Data/LTC1.Rdata")
+
+

### 2016-03-15 BPLE/LabTalk/LabTalk.tex
lab talk beamer file, plus slight changes to data prep. Will need to modify smoothing somehow to account for smaller counts in causes...

%% 
%%	This is file 'beamer_sample.tex'
%%	according to an MPIDR's PowerPoint template (?)
%%	
%%	by Eric Naujoks
%%
%%	Problems, bugs and comments to 
%%	naujoks@demogr.mpg.de
%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%	Praelegomena								%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%	- Make sure that you use utf8-encoding for all your .tex-files!!! (TeXnicCenter since version 2.0)
%%	- TeXnicCenter update: MPIDR intranet > Hard- & Sortfware > Software > Script and text editors > TeXnicCenter

\documentclass[20pt]{beamer}

\usepackage[ngerman,english]{babel}
\usepackage{graphicx}
\usepackage{tikz}
%\usepackage{animate} 
\usepackage[normalem]{ulem}
\geometry{paperwidth=10in, paperheight=7.5in}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{movie15}
%\usepackage{multimedia}
\usepackage[mpidr]{./mpidr/beamerthemeMPIDR}
\usepackage{color}
\definecolor{mygray}{rgb}{0.8,0.8,0.8}
%% Declaring title and author
\title{A low-mortality benchmark for the United States}
\subtitle{Tim Riffe, Adrien Remund, Magali Barbieri, Celeste Winant}		%%

%%	the institute's logo
\renewcommand{\mylogo}{\includegraphics[width=4.7in]{mpidr_logo_colour_en}}

%%	should be the very last package to be loaded
\usepackage{hyperref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%	Beginning of the document		%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%	titlepage - fixed frame:
%%	========================
\begin{frame}
	\titlepage
\end{frame}

%%	TOC:
%%	====
%\begin{frame}

%\end{frame}

% typical APC setup
\begin{frame}
\frametitle{A casual question}
\begin{quote}
How high would $e(0)$ be if we calculated it based on the rate schedule
composed of the lowest observed mortality from each cause in each age in a
set of populations?
\end{quote}
\color{mygray}
(Question due to happy hour chat)
\end{frame}

\begin{frame}
\frametitle{Expectations}
\begin{block}{prior I}
If life expectancy is high even large rate reductions don't translate to many
years gained (entropy).
\end{block}

\begin{block}{rule of thumb for current US mortality}
10\% reduction in rates buys about 1.5\% more $e(0)$\\
50\% reduction in rates buys about 10\% more $e(0)$
\end{block}
\color{mygray}
(So maybe this is no big deal?)
\end{frame}

\begin{frame}
\frametitle{Expectations}
\begin{block}{prior II}
More causes, more populations can only imply lower rates
\end{block}
\begin{block}{prior III}
Stochastic zeros don't imply no risk. We'll have to smooth rates to get rid of
them.
\end{block}
\end{frame}

\begin{frame}
\frametitle{Expectations}
\begin{block}{observation}
This exercise isn't very realistic, no matter how you spin it. But maybe it's
useful?
\end{block}
\end{frame}

\begin{frame}
\frametitle{Background}
\begin{block}{Not a new idea\ldots}
Vall\'{i}n \& Mesl{\'e} (2008) --- min age \& cause, countries\\
Wunsch (1975) --- min age, countries\\
Whelpton (1947) --- min age, countries
\end{block}
\end{frame}

\begin{frame}
\frametitle{Vall\'{i}n \& Mesl{\'e} (2008)}
\begin{figure}[b]
    \centering
      % figure made in R/APClab.R
    \includegraphics[scale=1.3]{Figures/Valline0.pdf}
    %\caption{Pertinent result from Vall\'{i}n \& Mesl{\'e} (2008)}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{Exercise}
\begin{block}{repeat for United States}
51 states (inlcuding DC)\\
homogenous data collection (no coding issues)\\
1959 - 2004 (2014 now)\\
12 (30) causes\\
ages 0-100+
\end{block}
\end{frame}

\begin{frame}
\frametitle{Males (1959-2004, 12 causes)}
\begin{figure}[b]
    \centering
    \includegraphics[scale=1.2]{Figures/e0trendsM.pdf}
    %\caption{Males (1959-2004\ldots)}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{Females (1959-2004, 12 causes)}
\begin{figure}[b]
    \centering
    \includegraphics[scale=1.2]{Figures/e0trendsF.pdf}
    %\caption{Females (1959-2004\ldots)}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{Why do I care?}
\begin{block}{I: Regular}
Low mortality benchmark is the straightest line. It rose a steady 0.18/yr for
males, 0.15/yr for females. USA $e(0)$ runs 30 years behind.
\end{block}
\begin{block}{II: Lazy interpretation}
Achievable via pure diffusion
(technology, health care access, risk factors) and no other kind of progress.
\end{block}
\begin{block}{III: Endogenous reference}
Record life expectancy (Japan\ldots) is used as a sign of the future.
Low mortality benchmark is a reference closer to home, difussion easier to imagine.
\end{block}
\end{frame}

\begin{frame}
\frametitle{Why do I care?}
\begin{block}{IV: Endogenous reference}
Seriously, some projection proposals use Japan as a guide\ldots
\end{block}
\begin{figure}[b]
    \centering
    \includegraphics[scale=.8]{Figures/Pascariu.png}
    %\caption{Females (1959-2004\ldots)}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{Why do I care?}
\begin{block}{IV: Endogenous reference}
Seriously, some projection proposals use Japan as a guide\ldots
\end{block}
\begin{figure}[b]
    \centering
    \includegraphics[scale=.8]{Figures/Medford.png}
    %\caption{Females (1959-2004\ldots)}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{Digression}
\begin{quote}
A feature of counterfactuals is that they aren't true.
\end{quote}
\end{frame}

\begin{frame}
\frametitle{Digression}
\begin{quote}
Period $e(0)$ is a counterfactual too, ergo \textbf{fake}, but useful.\\
Is it so bad to do fake things with the components of $e(0)$?
\end{quote}
\end{frame}

\begin{frame}
\frametitle{Digression}
\begin{quote}
After all, it is the case that \\
``If a cohort of individuals were subjected to
the mortality rate schedule implied by the low-mortality benchmark (minimum
aggregation), then the average length of life would be LMB $e(0)$.''
\end{quote}
\end{frame}

\begin{frame}
\frametitle{So why not do cool stuff with LMB?}
State gap in $e(0)$ due to shortcomings in cardiovascular diseases
\begin{figure}[b]
    \centering
    \includegraphics[scale=1.2]{Figures/Cardio.pdf}
    %\caption{Females (1959-2004\ldots)}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{So why not do cool stuff with LMB?}
Or: potential increase in $e(0)$ given lowest \textit{presently
achievable} $\mu^{cardio}$.
 \begin{figure}[b]
    \centering
    \includegraphics[scale=1.2]{Figures/Cardio.pdf}
    %\caption{Females (1959-2004\ldots)}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{What's next?}
\begin{itemize}
\item Redo with 30 causes, through 2014. AP-smoothing of causes is tougher with
30 causes because there are many zero counts. Practical tips there would be
nice.
\item We're leaning towards an easy-to-chew public health article rather than a
demographic methods spin. The method can be sufficiently expressed in
one or two sentences, nothing fancy here.
\item If it were a methods paper, might make sense to compare with a lifetable
eliminated of causes \textit{amenable to treatment}.
\item Poster at PAA (P1-85, early morning Thursday\ldots)
\end{itemize}
\end{frame}
%\begin{frame}
%\begin{itemize}
%  \item\item<1-> Chrono variation = mortality morbidity tradeoff
%  \item\item<2-> Thano variation = win
 % \item\item<3-> miscategorizing thano variation leads in wrong direction
%  \item\item<4-> assumptions were strong here\ldots
%\end{itemize}
%\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%	End of the document					%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2016-03-15 BPLE/R/DataPrep.R

@@ -109,7 +109,6 @@ sum(LT$Mx==0 & !is.na(LT$Mx))
 # this is 'hat' due to old age smoothing
 LT$Dxhat <- LT$Exposure * LT$mx
 
-
 # takes a while, warnings for interpolation (exppsure = 0)
 LT <- suppressWarnings(LT[, sm.chunk(.SD), by = list(State, Sex)])
 
@@ -122,28 +121,63 @@ LT$Age5                            <- LT$Age - LT$Age %% 5
 LT$Age5[LT$Age5 == 0 & LT$Age > 0] <- 1
 LT$Age5[LT$Age5 > 100]             <- 100 # because they'll all get same proportions...
 
-frac <- as.data.frame(frac)
-LT   <- as.data.frame(LT)
+#frac <- as.data.frame(frac)
+#LT   <- as.data.frame(LT)
 
+states <- unique(LT$State)
+st <- states[1]
+library(parallel)
+Nothing <- mclapply(states, function(st, LT){
+			ST      <- LT[LT$State == st, ]
+			fracc   <- local(get(load(paste0("Data/StateChunks/frac/",st,".Rdata"))))
+			STc 	<- do.call(rbind,lapply(unique(fracc$Cause), function(cause, ST){
+								ST$Cause <- cause
+								ST
+							}, ST = ST))
+			values        	<- fracc$Fxc
+			names(values) 	<- with(fracc,paste(State, Year, Sex, Age, Cause))
+			STc$cprop     	<- values[with(STc, paste(State, Year, Sex, Age5, Cause))]
+			# ok mxc is the product:
+			STc$mxc5 		<- STc$mxs * STc$cprop
+			STc$Dxc5 		<- STc$mxc5 * STc$Exposure
+			save(STc, file = paste0("Data/StateChunks/LTC/",st,".Rdata"))
+			NULL
+		}, LT=LT, mc.cores = 4)
+Nothing <- mclapply(states, function(st){
+		
+			STc   <- local(get(load(paste0("Data/StateChunks/LTC/",st,".Rdata"))))
+			STc   <- data.table(STc)
+			STc[, sm.chunk.cause(.SD), by = list(Sex, Cause)]
+			head(STc)
+			STc[, prop2 := makeprop(mxcs), by = list(Year, Sex, Age)]
+			STc$mxcs2 <- STc$prop2 * STc$mxs
[…]

### 2016-03-15 BPLE/R/e0trends.R

@@ -53,7 +53,7 @@ lines(1959:2004, apply(e0ms,2,max), col = "blue", lwd = 2, lty = 2)
 lines(1959:2004, bpm, col = "red",lwd=2)
 lines(1959:2004, use0m, col = "red",lwd=2,lty=2)
 legend("bottomright",lty=c(1,2,2,1),col=c("#00000050","red","blue","red"),
-		lwd=c(1,2,2,2),legend=c("state e(0)","US e(0)","vanguard e(0)","BP e(0)"),
+		lwd=c(1,2,2,2),legend=c("state e(0)","US e(0)","vanguard e(0)","LMB e(0)"),
 		bty="n")
 dev.off()
 
@@ -65,7 +65,7 @@ lines(1959:2004, apply(e0fs,2,max), col = "blue", lwd = 2, lty = 2)
 lines(1959:2004, bpf, col = "red",lwd=2)
 lines(1959:2004, use0f, col = "red",lwd=2,lty=2)
 legend("bottomright",lty=c(1,2,2,1),col=c("#00000050","red","blue","red"),
-		lwd=c(1,2,2,2),legend=c("state e(0)","US e(0)","vanguard e(0)","BP e(0)"),
+		lwd=c(1,2,2,2),legend=c("state e(0)","US e(0)","vanguard e(0)","LMB e(0)"),
 		bty="n")
 dev.off()
 

### 2016-03-15 BPLE/R/DataPrep.R
.

@@ -210,5 +210,3 @@ LTC$mxcs2 <- LTC$prop2 * LTC$mxs
 # so depending on whether the result is smooth or not, may be due for a second round of smoothing
 # within cause. Or maybe we accept this? Probably enough juice for now.
 save(LTC, file = "Data/LTC1.Rdata")
-
-
